---
title: "36-460/660 Final Project - Catcher Framing Heterogeneity"
author: "Joël Maïson"
date: 04/27/2026
format: pdf
---

Gen AI Usage Statement: This notebook was produced with the use of Claude (Anthropic) via the web chat interface at claude.ai. The purposes were for editing, commenting, paraphrasing, cleaning code and structure.

# ── Setup and Data Load ────────────────────────────────────────────────────

```{r load-libraries}
#| warning: false
#| message: false
library(tidyverse)
library(lme4)
library(broom.mixed)
library(knitr)
library(ggplot2)
library(forcats)
library(baseballr)
```

```{r load-cached-data}
# Raw Statcast data was pulled
# Loading the same RDS keeps the data source consistent across all three
# defensive components (framing, blocking, and this heterogeneity analysis).
sc_2025_raw <- readRDS("statcast_2025_raw.rds")
```

# ── Build Shadow Zone Dataset (with pitch_type preserved) ─────────────────

```{r shadow-threshold}
# Shadow zone threshold from MLB rulebook baseball circumference
# (9.000-9.250 inches), giving an average baseball width of ~0.242 ft.
# Source: https://www.mlb.com/glossary/statcast/catcher-framing
shadow_threshold <- ((9.000 / pi) + (9.250 / pi)) / 2 / 12
```

```{r build-shadow-data}
shadow_2025 <- sc_2025_raw %>%
  filter(
    description %in% c("called_strike", "ball"),
    !is.na(plate_x), !is.na(plate_z),
    !is.na(sz_top),  !is.na(sz_bot),
    !is.na(fielder_2),
    !is.na(pitch_type)
  ) %>%
  mutate(
    is_strike = as.integer(description == "called_strike"),
    pz_norm   = (plate_z - sz_bot) / (sz_top - sz_bot),
    dist_x    = pmin(abs(plate_x - 0.83), abs(plate_x + 0.83)),
    dist_zt   = abs(plate_z - sz_top),
    dist_zb   = abs(plate_z - sz_bot),
    dist_edge = pmin(dist_x, dist_zt, dist_zb),
    in_shadow = dist_edge <= shadow_threshold
  ) %>%
  filter(in_shadow) %>%
  mutate(
    catcher_id = as.factor(fielder_2),
    pitcher_id = as.factor(pitcher),
    batter_id  = as.factor(batter),
    count      = as.factor(paste0(balls, "-", strikes)),
    stand      = as.factor(stand),
    p_throws   = as.factor(p_throws),
    pitch_type = as.factor(pitch_type)
  ) %>%
  select(
    game_date, game_pk,
    pitcher_id, batter_id, catcher_id,
    is_strike, plate_x, plate_z, pz_norm,
    sz_top, sz_bot, dist_edge,
    count, stand, p_throws, pitch_type
  )
```

```{r build-shadow-model}
# 300 pitch minimum mirrors Morgan's framing model
eligible_catchers <- shadow_2025 %>%
  count(catcher_id) %>%
  filter(n >= 300) %>%
  pull(catcher_id)

shadow_model <- shadow_2025 %>%
  filter(catcher_id %in% eligible_catchers) %>%
  mutate(catcher_id = droplevels(catcher_id))

# Group rare pitch types into broader buckets so per-cell sample sizes stay healthy
shadow_model_pt <- shadow_model %>%
  mutate(
    pitch_type_grouped = case_when(
      pitch_type %in% c("FF", "FA", "FT", "SI") ~ "Fastball",
      pitch_type %in% c("SL", "ST", "SV")        ~ "Slider",
      pitch_type %in% c("CU", "KC", "CS")        ~ "Curveball",
      pitch_type == "CH"                          ~ "Changeup",
      pitch_type == "FC"                          ~ "Cutter",
      pitch_type == "FS"                          ~ "Splitter",
      TRUE                                        ~ "Other"
    ) %>% as.factor()
  )

cat("Pitches in modeling dataset:", nrow(shadow_model_pt), "\n")
cat("Eligible catchers:", n_distinct(shadow_model_pt$catcher_id), "\n")
shadow_model_pt %>% count(pitch_type_grouped, sort = TRUE)
```

# ── Exploratory Data Analysis ─────────────────────────────────────────────

```{r}
# ── EDA: Called strike rate by count and pitch type ────────────────────────
# Shows the raw heterogeneity in framing outcomes across two contextual
# dimensions before any modeling, motivating the count and pitch type
# interaction terms in the heterogeneity models.

eda_count_pt <- shadow_model_pt %>%
  group_by(count, pitch_type_grouped) %>%
  summarise(
    strike_rate = mean(is_strike),
    n_pitches   = n(),
    .groups     = "drop"
  ) %>%
  filter(n_pitches >= 50,
         pitch_type_grouped %in% c("Fastball", "Slider", "Curveball",
                                   "Changeup", "Cutter")) %>%
  mutate(
    count = factor(count, levels = c("0-0", "0-1", "0-2",
                                     "1-0", "1-1", "1-2",
                                     "2-0", "2-1", "2-2",
                                     "3-0", "3-1", "3-2"))
  )

eda_count_pt_plot <- eda_count_pt %>%
  ggplot(aes(x = count, y = strike_rate, color = pitch_type_grouped,
             group = pitch_type_grouped)) +
  geom_line(linewidth = 0.7) +
  geom_point(aes(size = n_pitches), alpha = 0.8) +
  scale_color_manual(
    values = c("Fastball"  = "firebrick",
               "Slider"    = "steelblue",
               "Curveball" = "darkgreen",
               "Changeup"  = "purple",
               "Cutter"    = "orange"),
    name = "Pitch type"
  ) +
  scale_size_continuous(name = "Pitches", labels = scales::comma) +
  scale_y_continuous(labels = scales::percent) +
  labs(
    title    = "Called strike rate by count and pitch type (2025 shadow zone)",
    subtitle = "Pitch type and count both shape outcomes on borderline pitches",
    x        = "Count",
    y        = "Called strike rate"
  ) +
  theme_bw() +
  theme(
    panel.grid.minor = element_blank(),
    axis.text.x      = element_text(angle = 45, hjust = 1),
    legend.position  = "bottom"
  )

eda_count_pt_plot
saveRDS(eda_count_pt_plot, "framing_eda_count_pt_plot.rds")
```
```{r}
# ── EDA: Called strike rate heatmap by count and pitch type ────────────────
# Distinct visual idiom from Morgan's count-by-handedness plot.
# Shows heterogeneity in framing outcomes across count and pitch type before
# any modeling, motivating the interaction terms.

eda_count_pt <- shadow_model_pt %>%
  group_by(count, pitch_type_grouped) %>%
  summarise(
    strike_rate = mean(is_strike),
    n_pitches   = n(),
    .groups     = "drop"
  ) %>%
  filter(n_pitches >= 50,
         pitch_type_grouped %in% c("Fastball", "Slider", "Curveball",
                                   "Changeup", "Cutter", "Splitter")) %>%
  mutate(
    count = factor(count, levels = c("0-0", "0-1", "0-2",
                                     "1-0", "1-1", "1-2",
                                     "2-0", "2-1", "2-2",
                                     "3-0", "3-1", "3-2")),
    pitch_type_grouped = factor(pitch_type_grouped,
                                levels = c("Fastball", "Cutter", "Slider",
                                           "Curveball", "Changeup", "Splitter"))
  )

eda_count_pt_plot <- eda_count_pt %>%
  ggplot(aes(x = count, y = pitch_type_grouped, fill = strike_rate)) +
  geom_tile(color = "white", linewidth = 0.4) +
  geom_text(aes(label = scales::percent(strike_rate, accuracy = 1)),
            size = 3, color = "black") +
  scale_fill_gradient2(
    low      = "dodgerblue",
    mid      = "white",
    high     = "firebrick",
    midpoint = 0.34,
    labels   = scales::percent,
    name     = "Called strike rate"
  ) +
  labs(
    title    = "Called strike rate by count and pitch type (2025 shadow zone)",
    subtitle = "Both contextual dimensions shape outcomes on borderline pitches",
    x        = "Count",
    y        = "Pitch type"
  ) +
  theme_bw() +
  theme(
    panel.grid       = element_blank(),
    axis.text.x      = element_text(angle = 0),
    legend.position  = "bottom"
  )

eda_count_pt_plot
saveRDS(eda_count_pt_plot, "framing_eda_count_pt_plot.rds")
```


# ── Fit Models (eval = FALSE; cached as RDS after first run) ───────────────

```{r fit-models, eval = FALSE}
# These four glmer fits each take 5 to 15 minutes. Fit once, save to RDS,
# then load below. Two reduced models are required because the count
# interaction model uses shadow_model (no pitch_type filter, 168,709 rows)
# while the pitch type interaction model uses shadow_model_pt (with
# pitch_type filter, 165,424 rows). LRTs require matched row counts.

# Count specific framing model: catcher-by-count interaction.
framing_count_model <- glmer(
  is_strike ~ plate_x + pz_norm + count + stand + p_throws +
              (1 | catcher_id) + (1 | pitcher_id) + (1 | batter_id) +
              (1 | catcher_id:count),
  data    = shadow_model,
  family  = binomial(link = "logit"),
  control = glmerControl(optimizer = "bobyqa", optCtrl = list(maxfun = 2e5))
)
saveRDS(framing_count_model, "framing_count_model.rds")

# Reduced model matched to the count model dataset (shadow_model)
framing_count_reduced <- glmer(
  is_strike ~ plate_x + pz_norm + count + stand + p_throws +
              (1 | catcher_id) + (1 | pitcher_id) + (1 | batter_id),
  data    = shadow_model,
  family  = binomial(link = "logit"),
  control = glmerControl(optimizer = "bobyqa", optCtrl = list(maxfun = 2e5))
)
saveRDS(framing_count_reduced, "framing_count_reduced.rds")

# Pitch type framing model: catcher-by-pitch-type interaction.
framing_pt_model <- glmer(
  is_strike ~ plate_x + pz_norm + count + stand + p_throws +
              (1 | catcher_id) + (1 | pitcher_id) + (1 | batter_id) +
              (1 | catcher_id:pitch_type_grouped),
  data    = shadow_model_pt,
  family  = binomial(link = "logit"),
  control = glmerControl(optimizer = "bobyqa", optCtrl = list(maxfun = 2e5))
)
saveRDS(framing_pt_model, "framing_pt_model.rds")

# Reduced model matched to the pitch type model dataset (shadow_model_pt)
framing_pt_reduced <- glmer(
  is_strike ~ plate_x + pz_norm + count + stand + p_throws +
              (1 | catcher_id) + (1 | pitcher_id) + (1 | batter_id),
  data    = shadow_model_pt,
  family  = binomial(link = "logit"),
  control = glmerControl(optimizer = "bobyqa", optCtrl = list(maxfun = 2e5))
)
saveRDS(framing_pt_reduced, "framing_pt_reduced.rds")
```

```{r load-models}
framing_count_model   <- readRDS("framing_count_model.rds")
framing_count_reduced <- readRDS("framing_count_reduced.rds")
framing_pt_model      <- readRDS("framing_pt_model.rds")
file.copy("framing_reduced.rds", "framing_pt_reduced.rds")
framing_pt_reduced    <- readRDS("framing_pt_reduced.rds")
```

# ── Likelihood Ratio Tests ─────────────────────────────────────────────────

```{r lrt-count}
# Tests whether the count interaction term earns its keep
lrt_count <- anova(framing_count_reduced, framing_count_model, test = "Chisq")
lrt_count
```

```{r lrt-pt}
# Tests whether the pitch type interaction term earns its keep
lrt_pt <- anova(framing_pt_reduced, framing_pt_model, test = "Chisq")
lrt_pt
```

```{r lrt-summary-table}
# Compact summary for the report
lrt_summary <- tibble(
  Model            = c("Count interaction", "Pitch type interaction"),
  `Chi-squared`    = c(round(lrt_count$Chisq[2],   2),
                       round(lrt_pt$Chisq[2],      2)),
  `Delta AIC`      = c(round(lrt_count$AIC[1] - lrt_count$AIC[2], 1),
                       round(lrt_pt$AIC[1]   - lrt_pt$AIC[2],     1)),
  `p-value`        = c(format.pval(lrt_count$`Pr(>Chisq)`[2], digits = 3),
                       format.pval(lrt_pt$`Pr(>Chisq)`[2],    digits = 3))
) %>%
  kable(
    format  = "simple",
    caption = "Likelihood ratio tests for framing heterogeneity"
  )

lrt_summary
```

# ── Catcher Names Lookup (cached) ──────────────────────────────────────────

```{r catcher-names}
if (file.exists("catcher_names.rds")) {
  catcher_names <- readRDS("catcher_names.rds")
} else {
  catcher_ids <- shadow_model_pt %>%
    distinct(catcher_id) %>%
    pull(catcher_id) %>%
    as.character() %>%
    as.numeric()

  catcher_names <- map_dfr(catcher_ids, function(id) {
    tryCatch({
      info <- mlb_people(person_ids = id)
      tibble(level = as.character(id), catcher_name = info$full_name)
    }, error = function(e) {
      tibble(level = as.character(id), catcher_name = as.character(id))
    })
  })
  saveRDS(catcher_names, "catcher_names.rds")
}
```

# ── Extract Random Effects with Conditional Standard Errors ────────────────

```{r extract-count-ranef}
# 90 percent CIs from conditional SEs follow Morgan's methodology
count_all_ranef <- tidy(framing_count_model, effects = "ran_vals")

catcher_main <- count_all_ranef %>%
  filter(group == "catcher_id") %>%
  arrange(estimate) %>%
  mutate(
    lower = estimate - 1.645 * std.error,
    upper = estimate + 1.645 * std.error,
    sig   = lower > 0 | upper < 0
  )

catcher_count <- count_all_ranef %>%
  filter(group == "catcher_id:count") %>%
  separate(level, into = c("catcher_id", "count"), sep = ":", remove = FALSE) %>%
  mutate(
    lower = estimate - 1.645 * std.error,
    upper = estimate + 1.645 * std.error,
    sig   = lower > 0 | upper < 0
  )
```

```{r extract-pt-ranef}
pt_ranef <- tidy(framing_pt_model, effects = "ran_vals") %>%
  filter(group == "catcher_id:pitch_type_grouped") %>%
  separate(level, into = c("catcher_id", "pitch_type_grouped"),
           sep = ":", remove = FALSE) %>%
  mutate(
    lower = estimate - 1.645 * std.error,
    upper = estimate + 1.645 * std.error,
    sig   = lower > 0 | upper < 0
  )
```

# ── Convert to Runs Saved Scale ────────────────────────────────────────────

```{r runs-saved-scale}
# Source for 0.125 runs per strike: https://www.mlb.com/glossary/statcast/catcher-framing
runs_per_strike <- 0.125

catcher_pitch_counts <- shadow_model_pt %>%
  count(catcher_id, name = "n_pitches") %>%
  mutate(catcher_id = as.character(catcher_id))

pt_pitch_counts <- shadow_model_pt %>%
  count(catcher_id, pitch_type_grouped, name = "n_pitches") %>%
  mutate(catcher_id = as.character(catcher_id))

catcher_count_runs <- catcher_count %>%
  left_join(catcher_pitch_counts, by = "catcher_id") %>%
  left_join(catcher_names, by = c("catcher_id" = "level")) %>%
  mutate(
    catcher_name  = ifelse(is.na(catcher_name), catcher_id, catcher_name),
    prob_estimate = plogis(estimate) - 0.5,
    prob_lower    = plogis(lower)    - 0.5,
    prob_upper    = plogis(upper)    - 0.5,
    runs_per100   = prob_estimate * 100 * runs_per_strike,
    runs_lower    = prob_lower    * 100 * runs_per_strike,
    runs_upper    = prob_upper    * 100 * runs_per_strike
  )

pt_runs <- pt_ranef %>%
  left_join(pt_pitch_counts, by = c("catcher_id", "pitch_type_grouped")) %>%
  left_join(catcher_names, by = c("catcher_id" = "level")) %>%
  mutate(
    catcher_name  = ifelse(is.na(catcher_name), catcher_id, catcher_name),
    prob_estimate = plogis(estimate) - 0.5,
    prob_lower    = plogis(lower)    - 0.5,
    prob_upper    = plogis(upper)    - 0.5,
    runs_per100   = prob_estimate * 100 * runs_per_strike,
    runs_lower    = prob_lower    * 100 * runs_per_strike,
    runs_upper    = prob_upper    * 100 * runs_per_strike
  )
```

# ── Variance Components Table ──────────────────────────────────────────────

```{r variance-table}
# Pull variance components from both models for the report
vc_count <- VarCorr(framing_count_model) %>% as.data.frame()
vc_pt    <- VarCorr(framing_pt_model)    %>% as.data.frame()

variance_components_table <- tibble(
  Model              = c(rep("Count interaction", nrow(vc_count)),
                         rep("Pitch type interaction", nrow(vc_pt))),
  `Random effect`    = c(vc_count$grp, vc_pt$grp),
  `Std. Dev.`        = round(c(vc_count$sdcor, vc_pt$sdcor), 3),
  Variance           = round(c(vc_count$vcov,  vc_pt$vcov),  4)
) %>%
  kable(
    format  = "simple",
    caption = "Variance components from heterogeneity models"
  )

variance_components_table
```

# ── Plot 1: Count Specific Caterpillar (catcher baselines) ────────────────

```{r plot-count-caterpillar}
catcher_main_plot <- catcher_main %>%
  mutate(level = fct_inorder(level)) %>%
  ggplot(aes(x = estimate, y = level, color = sig)) +
  geom_errorbarh(aes(xmin = lower, xmax = upper),
                 height = 0, alpha = 0.5, linewidth = 0.4) +
  geom_point(size = 1.5) +
  geom_vline(xintercept = 0, linetype = "dashed",
             color = "black", linewidth = 0.5) +
  scale_color_manual(
    values = c("FALSE" = "gray50", "TRUE" = "firebrick"),
    labels = c("FALSE" = "CI includes zero",
               "TRUE"  = "Significantly different from zero"),
    name   = NULL
  ) +
  labs(
    title    = "Catcher baseline framing effects (count specific model, 2025)",
    subtitle = "90% CI from conditional standard errors",
    x        = "Baseline framing random effect (log odds)",
    y        = NULL
  ) +
  theme_bw() +
  theme(
    axis.text.y        = element_blank(),
    axis.ticks.y       = element_blank(),
    panel.grid.major.y = element_blank(),
    panel.grid.minor   = element_blank(),
    legend.position    = "bottom"
  )

catcher_main_plot
```

# ── Plot 2: High Leverage Counts (0-2 and 3-0) ─────────────────────────────

```{r plot-leverage}
high_leverage <- catcher_count_runs %>%
  filter(count %in% c("0-2", "3-0")) %>%
  group_by(count) %>%
  arrange(desc(runs_per100)) %>%
  slice(c(1:10, (n() - 9):n())) %>%
  ungroup() %>%
  mutate(
    catcher_name = fct_reorder(catcher_name, runs_per100),
    direction    = ifelse(runs_per100 > 0, "Above average", "Below average")
  )

leverage_plot <- high_leverage %>%
  ggplot(aes(x = runs_per100, y = catcher_name, fill = direction)) +
  geom_col(width = 0.7, alpha = 0.85) +
  geom_errorbarh(aes(xmin = runs_lower, xmax = runs_upper),
                 height = 0.25, linewidth = 0.4, color = "gray30") +
  geom_vline(xintercept = 0, linewidth = 0.5, color = "black") +
  facet_wrap(~ count, scales = "free_y") +
  scale_fill_manual(
    values = c("Above average" = "steelblue",
               "Below average" = "firebrick"),
    name   = NULL
  ) +
  labs(
    title    = "Catcher framing in high leverage counts (2025)",
    subtitle = "Top and bottom 10 per count, runs saved per 100 shadow pitches",
    x        = "Runs saved per 100 pitches",
    y        = NULL
  ) +
  theme_bw() +
  theme(
    panel.grid.minor   = element_blank(),
    panel.grid.major.y = element_blank(),
    legend.position    = "bottom"
  )

leverage_plot
```

# ── Plot 3: Count Profiles (best and worst by baseline) ────────────────────

```{r plot-count-profiles}
profile_ids <- catcher_main %>%
  arrange(estimate) %>%
  slice(c(1:3, (n() - 2):n())) %>%
  pull(level) %>%
  as.character()

profile_data <- catcher_count_runs %>%
  filter(catcher_id %in% profile_ids) %>%
  mutate(
    count = factor(count, levels = c("0-0", "0-1", "0-2",
                                     "1-0", "1-1", "1-2",
                                     "2-0", "2-1", "2-2",
                                     "3-0", "3-1", "3-2"))
  )

profile_plot <- profile_data %>%
  ggplot(aes(x = count, y = runs_per100)) +
  geom_hline(yintercept = 0, linetype = "dashed", color = "black") +
  geom_errorbar(aes(ymin = runs_lower, ymax = runs_upper),
                width = 0.2, color = "gray40") +
  geom_point(aes(color = runs_per100 > 0), size = 2) +
  facet_wrap(~ catcher_name, ncol = 3) +
  scale_color_manual(values = c("TRUE" = "steelblue", "FALSE" = "firebrick"),
                     guide = "none") +
  labs(
    title    = "Catcher framing profiles by count (2025)",
    subtitle = "Three best and three worst catchers by baseline framing",
    x        = "Count",
    y        = "Runs saved per 100 pitches"
  ) +
  theme_bw() +
  theme(
    panel.grid.minor = element_blank(),
    axis.text.x      = element_text(angle = 45, hjust = 1)
  )

profile_plot
```

# ── Plot 4: Pitch Type Top and Bottom 5 ────────────────────────────────────

```{r plot-pt-topbot}
top_bot_by_pt <- pt_runs %>%
  filter(pitch_type_grouped %in% c("Fastball", "Slider", "Curveball", "Changeup")) %>%
  group_by(pitch_type_grouped) %>%
  arrange(desc(runs_per100)) %>%
  slice(c(1:5, (n() - 4):n())) %>%
  ungroup() %>%
  mutate(
    catcher_name = tidytext::reorder_within(catcher_name, runs_per100,
                                            pitch_type_grouped),
    direction    = ifelse(runs_per100 > 0, "Above average", "Below average")
  )

pt_topbot_plot <- top_bot_by_pt %>%
  ggplot(aes(x = runs_per100, y = catcher_name, fill = direction)) +
  geom_col(width = 0.7, alpha = 0.85) +
  geom_errorbarh(aes(xmin = runs_lower, xmax = runs_upper),
                 height = 0.25, linewidth = 0.4, color = "gray30") +
  geom_vline(xintercept = 0, linewidth = 0.5, color = "black") +
  facet_wrap(~ pitch_type_grouped, scales = "free_y", ncol = 2) +
  tidytext::scale_y_reordered() +
  scale_fill_manual(
    values = c("Above average" = "steelblue",
               "Below average" = "firebrick"),
    name   = NULL
  ) +
  labs(
    title    = "Catcher framing by pitch type (2025)",
    subtitle = "Top and bottom 5 per pitch type, runs saved per 100 shadow pitches",
    x        = "Runs saved per 100 pitches",
    y        = NULL
  ) +
  theme_bw() +
  theme(
    panel.grid.minor   = element_blank(),
    panel.grid.major.y = element_blank(),
    legend.position    = "bottom"
  )

pt_topbot_plot
```

# ── Plot 5: Pitch Type Specialist Profiles ─────────────────────────────────

```{r plot-pt-profiles}
profile_pt_ids <- pt_runs %>%
  group_by(catcher_id) %>%
  summarise(spread = max(runs_per100, na.rm = TRUE) -
                     min(runs_per100, na.rm = TRUE)) %>%
  arrange(desc(spread)) %>%
  slice(1:6) %>%
  pull(catcher_id)

profile_pt_data <- pt_runs %>%
  filter(catcher_id %in% profile_pt_ids) %>%
  mutate(
    pitch_type_grouped = factor(pitch_type_grouped,
                                levels = c("Fastball", "Cutter", "Slider",
                                           "Curveball", "Changeup",
                                           "Splitter", "Other"))
  )

pt_profile_plot <- profile_pt_data %>%
  ggplot(aes(x = pitch_type_grouped, y = runs_per100)) +
  geom_hline(yintercept = 0, linetype = "dashed", color = "black") +
  geom_errorbar(aes(ymin = runs_lower, ymax = runs_upper),
                width = 0.2, color = "gray40") +
  geom_point(aes(color = runs_per100 > 0), size = 2) +
  facet_wrap(~ catcher_name, ncol = 3) +
  scale_color_manual(values = c("TRUE" = "steelblue", "FALSE" = "firebrick"),
                     guide = "none") +
  labs(
    title    = "Catcher specialist profiles by pitch type (2025)",
    subtitle = "Six catchers with the largest spread across pitch types",
    x        = "Pitch type",
    y        = "Runs saved per 100 pitches"
  ) +
  theme_bw() +
  theme(
    panel.grid.minor = element_blank(),
    axis.text.x      = element_text(angle = 45, hjust = 1)
  )

pt_profile_plot
```

# ── Limitations: Why a Throwing Model Was Not Included ─────────────────────

```{r steals-diagnostic}
# The original project proposal included a throwing component (catcher arm
# effectiveness on stolen base attempts). The pitch level Statcast feed used
# for framing and blocking only tags terminal plate appearance events, so
# stolen base events do not appear in this dataset. The diagnostic below
# documents this limitation.

steals_check <- sc_2025_raw %>%
  filter(events %in% c("stolen_base_2b", "caught_stealing_2b",
                       "stolen_base_3b", "caught_stealing_3b",
                       "pickoff_caught_stealing_2b",
                       "pickoff_caught_stealing_3b")) %>%
  count(events)

cat("Stolen base events in pitch level feed: ", nrow(steals_check), " rows\n")
steals_check
```

# ── Save Outputs for report.qmd ────────────────────────────────────────────

```{r save-outputs}
saveRDS(catcher_main_plot,         "framing_main_plot.rds")
saveRDS(leverage_plot,             "framing_leverage_plot.rds")
saveRDS(profile_plot,              "framing_profile_plot.rds")
saveRDS(pt_topbot_plot,            "framing_pt_topbot_plot.rds")
saveRDS(pt_profile_plot,           "framing_pt_profile_plot.rds")
saveRDS(catcher_count_runs,        "framing_count_runs.rds")
saveRDS(pt_runs,                   "framing_pt_runs.rds")
saveRDS(lrt_summary,               "lrt_summary.rds")
saveRDS(variance_components_table, "variance_components_table.rds")

cat("Saved all framing heterogeneity outputs.\n")
```
