---
title: "Draft: Measuring MLB Catcher Defensive Contributions"
author: "Arthur Spirou, Joel Maison, and Morgan Corliss"
date: "15 April 2026"
format: 
  pdf:
    toc: false
execute:
  echo: false
  warning: false
  message: false
---

```{r}
#| echo: false
#| warning: false
#| message: false

library(tidyverse)
library(knitr)

# Load cached data
sc_2025_raw  <- readRDS("statcast_2025_raw.rds")

# Constants
shadow_threshold <- ((9.000 / pi) + (9.250 / pi)) / 2 / 12
runs_per_strike  <- 0.125

# Shadow zone filter
shadow_2025 <- sc_2025_raw %>%
  filter(
    description %in% c("called_strike", "ball"),
    !is.na(plate_x), !is.na(plate_z),
    !is.na(sz_top),  !is.na(sz_bot),
    !is.na(fielder_2)
  ) %>%
  mutate(
    is_strike = as.integer(description == "called_strike"),
    pz_norm   = (plate_z - sz_bot) / (sz_top - sz_bot),
    dist_x    = pmin(abs(plate_x - 0.83), abs(plate_x + 0.83)),
    dist_zt   = abs(plate_z - sz_top),
    dist_zb   = abs(plate_z - sz_bot),
    dist_edge = pmin(dist_x, dist_zt, dist_zb),
    in_shadow = dist_edge <= shadow_threshold
  ) %>%
  filter(in_shadow) %>%
  mutate(
    catcher_id = as.factor(fielder_2),
    pitcher_id = as.factor(pitcher),
    batter_id  = as.factor(batter),
    count      = as.factor(paste0(balls, "-", strikes)),
    stand      = as.factor(stand),
    p_throws   = as.factor(p_throws)
  ) %>%
  select(
    game_date, game_pk,
    pitcher_id, batter_id, catcher_id,
    is_strike, plate_x, plate_z, pz_norm,
    sz_top, sz_bot, dist_edge,
    count, stand, p_throws
  )

# Model dataset
eligible_catchers <- shadow_2025 %>%
  count(catcher_id) %>%
  filter(n >= 300) %>%
  pull(catcher_id)

shadow_model <- shadow_2025 %>%
  filter(catcher_id %in% eligible_catchers) %>%
  mutate(catcher_id = droplevels(catcher_id))

# Tables
model_comparison_table <- tibble(
  Model            = c("Full model", "Reduced model"),
  `Random effects` = c("Catcher + Pitcher + Batter", "Catcher only"),
  AIC              = c(204944.9, 205241.2),
  `Delta AIC`      = c(0.0, 296.3),
  `p-value`        = c("—", "< 2.2e-16")
) %>%
  kable(
    format  = "simple",
    caption = "Model comparison: full vs. reduced mixed-effects logistic regression",
    digits  = 1
  )

variance_fixed_table <- tibble(
  Term                     = c(
    "Pitcher (random)", "Batter (random)", "Catcher (random)",
    "plate_x (fixed)", "pz_norm (fixed)", "count 3-0 (fixed)",
    "count 0-2 (fixed)", "standR (fixed)", "p_throwsR (fixed)"
  ),
  Estimate                 = c(
    "—", "—", "—",
    "-0.014", "-0.047", "0.586", "-1.897", "0.140", "-0.012"
  ),
  `Std. Dev. / Std. Error` = c(
    "0.112", "0.126", "0.077",
    "0.006", "0.010", "0.038", "0.036", "0.016", "0.017"
  ),
  Variance                 = c(
    "0.013", "0.016", "0.006",
    "—", "—", "—", "—", "—", "—"
  ),
  `p-value`                = c(
    "—", "—", "—",
    "0.027", "< 2e-16", "< 2e-16", "< 2e-16", "< 2e-16", "0.468"
  )
) %>%
  kable(
    format  = "simple",
    caption = "Variance components and key fixed effects from full mixed-effects model",
    digits  = 3
  )

# Plot objects
heatmap_plot <- shadow_2025 %>%
  mutate(
    px_bin = round(plate_x / 0.1) * 0.1,
    pz_bin = round(plate_z / 0.1) * 0.1
  ) %>%
  filter(
    abs(px_bin) <= (0.83 + shadow_threshold + 0.05),
    pz_bin >= (1.5 - shadow_threshold - 0.05),
    pz_bin <= (3.5 + shadow_threshold + 0.05)
  ) %>%
  group_by(px_bin, pz_bin) %>%
  summarise(strike_rate = mean(is_strike), n_pitches = n(), .groups = "drop") %>%
  filter(n_pitches >= 50) %>%
  ggplot(aes(x = px_bin, y = pz_bin, fill = strike_rate)) +
  geom_tile() +
  annotate("rect", xmin = -0.83, xmax = 0.83, ymin = 1.5, ymax = 3.5,
           fill = NA, color = "black", linewidth = 0.8) +
  annotate("rect",
           xmin = -(0.83 + shadow_threshold), xmax = (0.83 + shadow_threshold),
           ymin = 1.5 - shadow_threshold, ymax = 3.5 + shadow_threshold,
           fill = NA, color = "white", linewidth = 0.6, linetype = "dashed") +
  scale_fill_gradient2(low = "dodgerblue", mid = "white", high = "firebrick",
                       midpoint = 0.5, name = "Called strike rate") +
  labs(
    title    = "Called strike rate by pitch location (2025 shadow zone)",
    subtitle = "Shadow zone only — interior excluded by design; dashed box = shadow zone boundary",
    x = "Horizontal position (ft, catcher's perspective)",
    y = "Vertical position (ft)"
  ) +
  coord_fixed(xlim = c(-1.5, 1.5), ylim = c(0.5, 4.5)) +
  theme_bw() +
  theme(panel.grid = element_blank())

pitchdist_plot <- shadow_2025 %>%
  count(catcher_id, name = "n_pitches") %>%
  ggplot(aes(x = n_pitches)) +
  geom_histogram(bins = 30, fill = "steelblue", color = "white") +
  geom_vline(xintercept = 300, color = "firebrick",
             linetype = "dashed", linewidth = 0.8) +
  annotate("text", x = 320, y = Inf, label = "300 pitch\nminimum",
           hjust = 0, vjust = 1.5, color = "firebrick", size = 3.5) +
  labs(
    title    = "Shadow zone pitches per catcher (2025 season)",
    subtitle = "111 total catchers; 85 meet the 300-pitch minimum",
    x = "Number of shadow zone pitches", y = "Number of catchers"
  ) +
  theme_bw() +
  theme(panel.grid.minor = element_blank())

countrate_plot <- shadow_2025 %>%
  group_by(count) %>%
  summarise(strike_rate = mean(is_strike), n_pitches = n(), .groups = "drop") %>%
  mutate(count = fct_reorder(count, strike_rate)) %>%
  ggplot(aes(x = strike_rate, y = count, fill = strike_rate)) +
  geom_col() +
  geom_text(aes(label = scales::percent(strike_rate, accuracy = 0.1)),
            hjust = -0.1, size = 3) +
  scale_fill_gradient2(low = "dodgerblue", mid = "white", high = "firebrick",
                       midpoint = mean(shadow_2025$is_strike), guide = "none") +
  scale_x_continuous(labels = scales::percent,
                     expand = expansion(mult = c(0, 0.12))) +
  labs(
    title    = "Called strike rate by count in shadow zone (2025 season)",
    subtitle = "Pitcher-friendly counts show higher strike rates on borderline pitches",
    x = "Called strike rate", y = "Count"
  ) +
  theme_bw() +
  theme(panel.grid.minor = element_blank())

handedness_plot <- shadow_2025 %>%
  group_by(stand, count) %>%
  summarise(strike_rate = mean(is_strike), n_pitches = n(), .groups = "drop") %>%
  mutate(
    count = fct_reorder(count, strike_rate),
    stand = recode(stand, "L" = "Left-handed batter", "R" = "Right-handed batter")
  ) %>%
  ggplot(aes(x = strike_rate, y = count, color = stand)) +
  geom_point(aes(size = n_pitches), alpha = 0.7) +
  geom_line(aes(group = count), color = "gray70", linewidth = 0.4) +
  scale_color_manual(
    values = c("Left-handed batter" = "steelblue", "Right-handed batter" = "firebrick"),
    name = NULL
  ) +
  scale_size_continuous(name = "Pitches", range = c(2, 6),
                        labels = scales::comma) +
  scale_x_continuous(labels = scales::percent) +
  labs(
    title    = "Called strike rate by count and batter handedness (2025 shadow zone)",
    subtitle = "Lines connect the same count across handedness — gap indicates handedness effect",
    x = "Called strike rate", y = "Count"
  ) +
  theme_bw() +
  theme(panel.grid.minor = element_blank(), legend.position = "bottom")
```

# Abstract

# Introduction

In baseball, catchers play a pivotal role on defense, yet their value is 
historically difficult to measure. Traditional statistics such as passed balls 
and caught stealing rate capture only a fraction of what a catcher contributes. 
In this project we focus on three key aspects of catcher defense: pitch framing, 
pitch blocking, and caught stealing prevention. We then quantified each in terms of 
runs saved, an easy to understand metric that allows for direct comparison across
defensive skills and is accessible to even the most casual baseball fan.

Pitch framing, the ability to influence umpires to call borderline pitches 
strikes, is one of the most impactful and measurable aspects of catcher defense. 
A single extra strike per game can shift run expectancy, heighten pitcher 
confidence, and alter game outcomes. Elite framers can provide significant value,
in the 2024 MLB season, the top-rated framing catcher Patrick Bailey saved 22 runs
through framing alone, with the next five catchers saving between 11 and 13 runs
(MLB Advanced Media, 2025). With the rise of pitch tracking, pitch coordinates 
are now widely available, framing is another way teams can evaluate catchers to 
ensure their getting the best guys for the job. 

To measure framing, we fit a mixed-effects logistic regression model to 
168,709 shadow zone pitches from the 2025 MLB season, isolating each catcher's 
framing contribution as a random effect after controlling for pitch location, 
count, batter, and pitcher. Results reveal meaningful variation in framing 
ability across MLB catchers. [Pending bootstrap completion — uncertainty 
estimates and catcher rankings to be added prior to final submission.

Blocking potentially wild pitches, swinging strikes, and balls that are low and/or off the plate is an extremely important defensive trait for catchers. This is especially true when runners are on base, because successfully blocking a ball can prevent immediate runs or runner advancement on a given pitch while also shaping the rest of the inning by keeping runners from moving into more advantageous positions. Traditionally, Baseball Savant has used advanced inputs such as catcher positioning and called pitch type, which are not available in public pitch-level data, to build catcher blocking metrics and rankings. We aim to develop a comparable model for ranking catcher blocking ability using only publicly available pitch-level data.

Blocking pitches has effects that go beyond the box score. In addition to directly preventing passed balls, wild pitches, runs, and stolen bases, strong blocking can also influence pitcher and team confidence. When a pitcher trusts their catcher to handle balls in the dirt or pitches that break well outside the strike zone, they are better able to use their full arsenal aggressively against hitters. Many of the most effective pitches are designed to look hittable before finishing in locations that are difficult or impossible to square up, often below the zone or well off the plate. In high-leverage situations against strong hitters, the ability to throw those pitches with confidence can have a major impact on game outcomes and on a pitcher’s overall effectiveness. Because many catchers are valued more for their defense than their offense, methods for evaluating pitch-blocking skill are especially important for teams. Quantifying blocking ability can improve catcher evaluation on its own, while also supporting better analysis of pitchers and providing teams with additional insight into the effectiveness of specific pitcher-catcher battery combinations.

# Data

Pitch-level Statcast data was accessed via Baseball Savant's CSV export API for 
the 2025 MLB regular season (March 27 – September 28, 2025), yielding 712,192 
total pitches across 118 variables (MLB Advanced Media, 2025). Key variables 
include `plate_x` and `plate_z` (pitch coordinates at home plate), `sz_top` and 
`sz_bot` (batter-specific strike zone boundaries), `fielder_2` (catcher ID), 
`description` (called strike or ball), `balls`, `strikes`, `stand`, and 
`p_throws`. Pitch-level coordinate data is uniquely suited for framing analysis 
as it allows precise separation of where the pitch actually crossed the plate 
from how the catcher influenced the umpire's call.

Key variables for the blocking analysis: plate_x, plate_z (pitch coordinates), sz_top, sz_bot (batter-specific strike zone boundaries), fielder_2 (catcher ID),    pitch_type, release_speed, description (including blocked_ball and swinging_strike_blocked), balls, strikes, p_throws, and des (used to identify official wild     pitch / passed ball outcomes for a secondary check)

## Preprocessing: Catcher Framing

The dataset was filtered to called pitches only. Swings, fouls, and balls in 
play were excluded since framing only applies to pitches the umpire must judge. 
Pitches with missing plate coordinates, strike zone boundaries, or catcher ID 
were removed. A shadow zone filter was then applied, retaining only pitches 
within one baseball diameter of the strike zone edge, consistent with MLB's 
official shadow zone definition (MLB.com Glossary). The threshold of 0.242 ft 
was derived from the 2026 Official Baseball Rules circumference specification of 
9.000–9.250 inches. Vertical pitch location was normalized (`pz_norm`) relative 
to each batter's individual strike zone to control for batter height differences. 
Finally, a minimum of 300 shadow zone pitches per catcher was applied to ensure 
reliable random effect estimates, reducing the dataset from 111 to 85 catchers 
and from 171,304 to 168,709 pitches. The final model dataset contains 168,709 
shadow zone pitches across 85 catchers.

## Preprocessing: Catcher Blocking

We filtered the data to pitches with non-missing values for plate_x, plate_z, sz_top, sz_bot, fielder_2, pitch_type, and release_speed. We then defined blocking opportunities as pitches below the strike zone, where plate_z < sz_bot, to isolate pitches on which a block was plausibly required. A binary outcome variable, is_block, was created and set equal to 1 when the pitch description was blocked_ball or swinging_strike_blocked, and 0 otherwise, allowing us to define block rate as is_block over blocking opportunities. We also renamed fielder_2 and pitcher to catcher_id and pitcher_id, created a count variable from the balls-strikes combination, and formatted pitch_type and p_throws as categorical variables. To avoid unstable coefficient estimates, pitch types with fewer than 500 below-zone pitches were grouped into an "Other" category. We considered including a swing indicator, but removed it because swinging_strike_blocked already implies a recorded block. The resulting simple blocking model dataset contains 157,195 below-zone pitches across 96 catchers and 857 pitchers. Finally, we used a catcher-inclusion threshold of 100 below-zone pitches, since blocking opportunities are much rarer than framing opportunities.

## Exploratory Data Analysis: Catcher Framing

The strike rate for pitches in the shadow zone is 34.2%. This confirms that 
borderline pitches are genuinely ambiguous, umpires do not consistently call them
strikes. A heatmap of called strike rate by pitch location reveals a clear gradient
from near-certain strikes at the interior edge of the shadow zone to near-certain 
balls at the exterior edge, confirming that borderline calls are not deterministic
(@fig-heatmap). The interior of the strike zone is excluded by design since pitches 
well inside the zone are not influenced by framing.

@fig-pitchdist displays the distribution of shadow zone pitches per catcher. This
figure is right-skewed with a long right tail of starting catchers. The 
300-pitch minimum cutoff removes 26 catchers with insufficient data, retaining 
85 catchers with reliable sample sizes for modeling. Catchers retained for 
modeling saw anywhere from 300 to approximately 4,500 shadow zone pitches.

@fig-countrate shows that called strike rate varies dramatically by count, ranging 
from 57% on 3-0 counts to 10.6% on 0-2 counts. This confirms that count 
strongly influences umpire behavior on borderline pitches. Notably, the highest 
strike rates occur on hitter-friendly counts like 3-0 and 2-0, where umpires 
appear more willing to call borderline pitches strikes when the pitcher needs 
to throw one, while two-strike counts like 0-2 and 1-2 show the lowest strike 
rates. This pattern justifies inclusion of `count` as a fixed effect in the 
model to avoid attributing count effects to catcher framing.

Finally, @fig-handedness reveals that right-handed batters consistently receive higher
called strike rates than left-handed batters across every count with the largest
gap observed on 3-1 counts. This persistent handedness effect justifies including
`stand` as a fixed effect to avoid attributing batter handedness differences to
catcher framing.

## Exploratory Data Analysis: Catcher Blocking

Inspection of the description field confirmed that Statcast includes explicit labels for successful blocks through blocked_ball and swinging_strike_blocked. By contrast, official blocking failures appear extremely sparse in publicly available pitch-level data: searching the des field identified only 86 wild pitches and 29 passed balls. This strong imbalance suggests that official wild pitch and passed ball outcomes substantially undercount true blocking failures, since many missed or difficult blocks are not recorded as official failures. For example, a passed ball with no runners on base may simply be scored as a ball rather than as a passed ball. Because of this limitation, recorded block rate among below-zone opportunities will serve as the primary modeling outcome, while official wild pitch and passed ball outcomes will be used only in a secondary supplemental model for evaluation. Blocking opportunities are also distributed unevenly across catchers, with a long tail of low-volume catchers, so a 100-pitch cutoff helps stabilize catcher-level random-effect estimates. In addition, there is substantial variation in below-zone opportunity counts across pitch types, which is why sparse pitch types are collapsed into an Other category before model fitting.


# Methods

## Morgan Corliss: Catcher Framing

The goal of this analysis is to isolate each catcher's influence on borderline 
pitch calls above and beyond what can be explained by pitch location, count, and 
batter/pitcher context. To accomplish this, we fit a mixed-effects logistic 
regression model where the binary outcome is whether a shadow zone pitch was 
called a strike (1) or a ball (0). The model takes the following form:

$$\log\left(\frac{p_i}{1-p_i}\right) = \beta_0 + \beta_1 \text{plate\_x}_i + 
\beta_2 \text{pz\_norm}_i + \beta_3 \text{count}_i + \beta_4 \text{stand}_i + 
\beta_5 \text{p\_throws}_i + u_{\text{catcher}[i]} + u_{\text{pitcher}[i]} + 
u_{\text{batter}[i]}$$

where $p_i$ is the probability of a called strike on pitch $i$. The fixed effects 
control for pitch location (`plate_x`, `pz_norm`), count, batter handedness 
(`stand`), and pitcher handedness (`p_throws`). The random effects 
$u_{\text{catcher}[i]}$, $u_{\text{pitcher}[i]}$, and $u_{\text{batter}[i]}$ 
capture the influence of each catcher, pitcher, and batter respectively. The 
catcher random effect is the primary quantity of interest — it represents each 
catcher's tendency to influence umpire calls above or below the league average 
after accounting for all other factors.

The model assumes the binary outcome follows a Bernoulli distribution: 
$Y_i \sim \text{Bernoulli}(p_i)$, where the log-odds of a called strike are 
modeled as a linear function of the fixed and random effects. The catcher, 
pitcher, and batter random effects are each assumed to be independently normally 
distributed with mean zero: $u_j \sim N(0, \sigma^2)$. Random effects are 
assumed independent of each other and of the fixed effect predictors, and shadow 
zone pitches are assumed conditionally independent given the random effects.

A mixed-effects model was chosen over a fixed-effects specification for several 
reasons. With 85 catchers in the dataset, a fixed-effects approach would require 
estimating 84 catcher indicator coefficients, producing unreliable estimates for 
backup catchers with small sample sizes. The mixed-effects framework handles this 
naturally through partial pooling — shrinking estimates for catchers with limited 
data toward the league mean while allowing well-sampled catchers to deviate 
further. Pitcher and batter random effects were included to avoid attributing 
their tendencies to catchers, which would otherwise bias the framing estimates. 
The BOBYQA optimizer was specified following recommendations for mixed models 
with complex random effect structures (Powell, 2009).

To evaluate the model, we fit a reduced specification excluding the pitcher and 
batter random effects:

$$\text{is\_strike} \sim \text{plate\_x} + \text{pz\_norm} + \text{count} + 
\text{stand} + \text{p\_throws} + (1 | \text{catcher\_id})$$

The full and reduced models were compared via AIC, where a lower value indicates 
better fit relative to model complexity. A likelihood ratio test was also 
conducted to formally assess whether adding pitcher and batter random effects 
significantly improves model fit. If the full model fits better, it confirms 
that omitting pitcher and batter effects would bias the catcher framing estimates.

To quantify uncertainty in the catcher random effect estimates, we employ a 
structured bootstrap... [having trouble with bootstrap, not complete at this time]

## Arthur Spirou: Catcher Blocking via Mixed-Effects Logistic Regression

### Blocking Model Description**
- Goal: isolate catcher influence on likely blocking opportunities above and beyond pitch location, pitch type, and pitcher context
- Mixed-effects logistic regression with binary outcome: recorded block (1) or no recorded block (0)
- Primary outcome defined as is_block = 1 when description was blocked_ball or swinging_strike_blocked
- Restricted to below-zone pitches (plate_z < sz_bot)
- Simple model fixed effects: plate_x, plate_z, and collapsed pitch type
- Simple model random effects: catcher_id, pitcher_id
- Full model will build on the simple model to include release_speed, count, and p_throws
- Catcher random effect represents each catcher’s ability to convert likely blocking opportunities into recorded blocks, relative to league average. 
  Controlling for block attempt difficulty and pitcher

### Blocking Model Assumptions
- Binary outcome follows a Bernoulli distribution
- log-odds of a recorded block modeled as a linear function of fixed and random effects
- catcher and pitcher random effects are assumed to be independent. Also assumed to
  be normally distributed with mean zero
- random effects are independent of fixed effects
- The below-zone filter is assumed to capture a meaningful set of likely blocking opportunities,
  however it does not perfectly identify every true blocking situation
- Outcome is interpreted as a recorded block rate, not as a fully observed success-versus-failure blocking label

Simple Blocking Model Formulation:

$$
Y_i \mid u_{j(i)}, v_{k(i)} \sim \mathrm{Bernoulli}(p_i)
$$

$$
\log\left(\frac{p_i}{1 - p_i}\right)
=
\beta_0
+ \beta_1 \,\mathrm{plate\_x}_i
+ \beta_2 \,\mathrm{plate\_z}_i
+ \boldsymbol{\gamma}^{\top}\mathbf{z}_i
+ u_{j(i)}
+ v_{k(i)}
$$

$$
u_j \sim N(0, \sigma^2_{\mathrm{catcher}})
$$

$$
v_k \sim N(0, \sigma^2_{\mathrm{pitcher}})
$$

Here, $Y_i$ indicates whether below-zone pitch $i$ resulted in a recorded block, $\mathbf{z}_i$ is the vector of indicator variables for `pitch_type_collapsed`, $u_{j(i)}$ is the catcher-specific random intercept, and $v_{k(i)}$ is the pitcher-specific random intercept.

### Blocking Model Justification
- Mixed-effects logistic regression is appropriate because the outcome is binary. 
  Also, repeated observations from the same catchers and pitchers create dependence
- Treating catcher as a random effect allows partial pooling, which stabilizes estimates 
  for lower-volume catchers and avoids overfitting
- Pitcher randomeffects are particularly crucial and meaningful for a blocking model because 
  some pitchers throw balls in the dirt or balls that move dramatically away from the strike zone
  more often than others. This can both create more blocking opportunities and pose more challenging
  blocking opportunities for certain catchers.
- Similar to pitch framing model, BOBYQA optimizer specified due to large dataset size and crossed random effects nature

### Blocking Model Comparison/Evaluation
Simple Model: is_block ~ plate_x + plate_z + pitch_type_collapsed + (1 | catcher_id) + (1 | pitcher_id)
Full Candidate Model: is_block ~ plate_x + plate_z + pitch_type_collapsed + release_speed + count + p_throws + (1 | catcher_id) + (1 | pitcher_id)
- Plan to compare the simple and full models using AIC, where lower AIC indicates 
  better out-of-sample tradeoff between fit and complexity
- A likelihood ratio test will be used to assess whether the additional fixed effects in the 
  full model significantly improve fit
- Catcher effects will be ranked and compared to Baseball Savant's catcher blocking ratings from the 2025 season

### Blocking Model Uncertainty Quantification
- Similar to the framing model, we plan to use structured bootstrap resampling with the unit of resampling being
  plate appearances within games to preserve within-game depe ndence and catcher usage patterns
- Each bootstrap sample will refit the selected blocking model and store the catcher random effects
- Plan to report 90% confidence intervals for catcher blocking effects. Random seed will be used for reproducibility.

# Results

## Morgan Corliss: Catcher Framing

### Model Comparison

The full model achieved a substantially lower AIC (204,945) compared to the 
reduced model (205,241), a difference of 296 units — well above the conventional 
threshold of 2. The likelihood ratio test confirms this improvement is 
statistically significant (χ²(2) = 300.25, p < 2.2e-16), providing strong 
evidence that pitcher and batter random effects meaningfully improve model fit. 
This result confirms that pitcher and batter tendencies independently influence 
borderline pitch calls above and beyond pitch location and count — omitting them 
would attribute some of their influence to catchers, biasing the framing 
estimates. The full model was therefore selected for all subsequent inference on 
catcher framing effects. These results are summarized in (@tbl-model-comparison).

### Model Interpretation

@tbl-variance-fixed displays the variance components and key fixed effects from the full 
model. Catcher variance (0.006) is smaller than both pitcher (0.013) and batter 
(0.016) variance, suggesting that pitch location and batter/pitcher context 
explain more variation in borderline calls than catcher framing alone. However, 
catcher variance is meaningfully non-zero, confirming that framing ability varies 
across catchers above and beyond these controls and justifying the framing 
analysis.

Among the fixed effects, pitcher handedness (`p_throwsR`) was not statistically 
significant (β = -0.012, p = 0.468), suggesting that after controlling for pitch 
location and count, pitcher handedness does not meaningfully influence borderline 
call rates. It was retained in the model for completeness. All count coefficients 
were highly significant (p < 2e-16), consistent with the dramatic variation in 
called strike rates across counts observed in the EDA — with 3-0 counts showing 
the highest tendency toward called strikes (β = 0.586) and 0-2 counts the lowest 
(β = -1.897). Right-handed batters received significantly more called strikes on 
borderline pitches than left-handed batters (β = 0.140, p < 2e-16), consistent 
with the handedness effect observed in (@fig-handedness).

### Catcher Framing Estimates

[Pending bootstrap completion — caterpillar plot, ridge plot, runs saved 
leaderboard, and uncertainty estimates to be added prior to final submission]

## Arthur Spirou: Catcher Blocking Results

The simple blocking model converged successfully on 157,195 below-zone pitches drawn from 96 catchers and 857 pitchers. Both catcher- and pitcher-level random-effect variances were significant, indicating that blocking outcomes vary across both groups even after controlling for pitch type and pitch location. Notably, the pitcher-level variance (0.059) exceeded the catcher-level variance (0.022), suggesting that pitcher tendencies and overall pitch difficulty explain more variation in recorded blocks than catcher differences alone. In the simple model, vertical pitch location (plate_z) was the dominant predictor of a recorded block (estimate = -2.218, p < 2e-16), meaning that lower pitches were much more likely to result in a block. Horizontal pitch location (plate_x) was also a significant predictor (estimate = -0.061, p < 0.001). In addition, different types of breaking balls were associated with higher odds of a recorded block, while four-seam fastballs and sinkers had the lowest odds. The full model will be compared with the simple model using AIC and a likelihood ratio test. Finally, official wild pitch and passed ball labels were found to be extremely sparse, with only 86 wild pitches and 29 passed balls identified, further reinforcing the decision not to use official failures as the primary blocking outcome.

# Discussion

## Conclusions

The model comparison results provide strong evidence that catcher framing is a 
real and measurable skill. The full mixed-effects model fit significantly better 
than the reduced model (AIC difference = 296, χ²(2) = 300.25, p < 2.2e-16), 
confirming that catcher identity explains meaningful variation in borderline 
pitch calls above and beyond pitch location, count, pitcher, and batter effects. 
The non-zero catcher variance (0.006) further supports this conclusion. 

*[Pending bootstrap completion — conclusions about the magnitude of framing 
differences across catchers and runs saved estimates to be added prior to 
final submission.]*

## Limitations

Several limitations of this analysis should be noted. First, the shadow zone 
threshold of 0.242 ft is derived from the average baseball diameter per the 
2026 Official Baseball Rules and a robustness check with alternative values 
would strengthen this analysis. Second, this analysis is based on a single 
season of data, framing estimates from one season may reflect some noise. 
Conducting an analysis of different years would help verify our conclusions.

The blocking analysis has several important limitations. First, the blocking outcome is not a perfect success-versus-failure label: the 0 class includes below-zone pitches that did not receive a recorded block, but some of those pitches may have been received cleanly rather than missed. Official wild pitch and passed ball outcomes are too sparse to serve as the primary response variable, and they likely undercount true blocking failures because many difficult or mishandled pitches are never scored as official failures. In addition, the below-zone filter (plate_z < sz_bot) is only a practical approximation of blocking opportunities. While it helps isolate pitches where a block is plausibly required, it does not account for pitches far above the zone, may exclude some horizontally difficult pitches, and may still include low pitches that were relatively easy for the catcher to receive. The model is also limited to publicly available pitch-level variables and therefore cannot incorporate features such as catcher starting position, body orientation, glove-side versus backhand mechanics, or other tracking information used in proprietary blocking models like Baseball Savant's. Finally, pitch type was collapsed for sparse categories to improve model stability, but doing so may hide meaningful differences among rare pitch types, such as an eephus pitch.

## Future Work

Several extensions of this work would strengthen the conclusions. A 
year-to-year stability check using 2024 data would help determine 
whether catcher rankings are consistent across seasons. High correlation between 
2024 and 2025 estimates would provide evidence that the model is measuring real 
framing skill rather than noise. Finally, a Bayesian implementation using 
`rstanarm` would provide full posterior distributions on catcher framing effects
rather than bootstrap approximations, accounting for the hierarchical structure
of the data.

# References

MLB.com. "Catcher Framing." *MLB Glossary*, Major League Baseball, 
https://www.mlb.com/glossary/statcast/catcher-framing. Accessed April 2026.

MLB Advanced Media. "Catcher Framing Leaderboard." *Baseball Savant*, 
Major League Baseball, 2025. 
https://baseballsavant.mlb.com/catcher_framing. Accessed April 2026.

MLB Advanced Media. *Statcast Pitch-Level Data, 2025 MLB Regular Season.* 
Baseball Savant, Major League Baseball, 2025. 
https://baseballsavant.mlb.com/statcast_search. Accessed April 2026.

Major League Baseball. *2026 Official Baseball Rules*, pg. 5. 
https://mktg.mlbstatic.com/mlb/official-information/2026-official-baseball-rules.pdf. 
Accessed April 2026.

Powell, M.J.D. "The BOBYQA Algorithm for Bound Constrained Optimization Without 
Derivatives." *Cambridge NA Report NA2009/06*, University of Cambridge, 2009.

# Appendix

## Tables

```{r}
#| echo: false
#| label: tbl-model-comparison
#| tbl-cap: "Table 1: Model comparison — full vs. reduced mixed-effects logistic regression"

model_comparison_table
```

```{r}
#| echo: false
#| label: tbl-variance-fixed
#| tbl-cap: "Table 2: Variance components and key fixed effects from full mixed-effects model"

variance_fixed_table
```

## Figures

```{r}
#| echo: false
#| warning: false
#| fig-cap: "Figure 1: Called strike rate by pitch location (2025 shadow zone). Solid box = strike zone, dashed box = shadow zone boundary. Interior of strike zone excluded by design."
#| label: fig-heatmap
#| fig-width: 6
#| fig-height: 5

heatmap_plot
```

```{r}
#| echo: false
#| warning: false
#| fig-cap: "Figure 2: Distribution of shadow zone pitches per catcher (2025 season). Dashed line indicates 300-pitch minimum threshold."
#| label: fig-pitchdist
#| fig-width: 6
#| fig-height: 4

pitchdist_plot
```

```{r}
#| echo: false
#| warning: false
#| fig-cap: "Figure 3: Called strike rate by count in shadow zone (2025 season)."
#| label: fig-countrate
#| fig-width: 6
#| fig-height: 5

countrate_plot
```

```{r}
#| echo: false
#| warning: false
#| fig-cap: "Figure 4: Called strike rate by count and batter handedness (2025 shadow zone)."
#| label: fig-handedness
#| fig-width: 6
#| fig-height: 5

handedness_plot
```

