---
title: "Draft: Measuring MLB Catcher Defensive Contributions"
author: "Arthur Spirou, Joel Maison, and Morgan Corliss"
date: "15 April 2026"
format: 
  pdf:
    toc: false
execute:
  echo: false
  warning: false
  message: false
---

```{r}
#| echo: false
#| warning: false
#| message: false

# Load libraries
library(httr)
library(tidyverse)
library(lme4)
library(broom.mixed)
library(ggridges)

# Load cached data
sc_2025_raw <- readRDS("statcast_2025_raw.rds")
#bootstrap_framing <- readRDS("bootstrap_framing_50.rds")
```

# Abstract

# Introduction

In baseball, catchers play a pivotal role on defense, yet their value is 
historically difficult to measure. Traditional statistics such as passed balls 
and caught stealing rate capture only a fraction of what a catcher contributes. 
In this project we focus on three key aspects of catcher defense: pitch framing, 
pitch blocking, and caught stealing prevention. We then quantified each in terms of 
runs saved, an easy to understand metric that allows for direct comparison across
defensive skills and is accessible to even the most casual baseball fan.

Pitch framing, the ability to influence umpires to call borderline pitches 
strikes, is one of the most impactful and measurable aspects of catcher defense. 
A single extra strike per game can shift run expectancy, heighten pitcher 
confidence, and alter game outcomes. Elite framers can provide significant value,
in the 2024 MLB season, the top-rated framing catcher Patrick Bailey saved 22 runs
through framing alone, with the next five catchers saving between 11 and 13 runs
(MLB Advanced Media, 2025). With the rise of pitch tracking, pitch coordinates 
are now widely available, framing is another way teams can evaluate catchers to 
ensure their getting the best guys for the job. 

To measure framing, we fit a mixed-effects logistic regression model to 
168,709 shadow zone pitches from the 2025 MLB season, isolating each catcher's 
framing contribution as a random effect after controlling for pitch location, 
count, batter, and pitcher. Results reveal meaningful variation in framing 
ability across MLB catchers. [Pending bootstrap completion — uncertainty 
estimates and catcher rankings to be added prior to final submission.]

# Data

Pitch-level Statcast data was accessed via Baseball Savant's CSV export API for 
the 2025 MLB regular season (March 27 – September 28, 2025), yielding 712,192 
total pitches across 118 variables (MLB Advanced Media, 2025). Key variables 
include `plate_x` and `plate_z` (pitch coordinates at home plate), `sz_top` and 
`sz_bot` (batter-specific strike zone boundaries), `fielder_2` (catcher ID), 
`description` (called strike or ball), `balls`, `strikes`, `stand`, and 
`p_throws`. Pitch-level coordinate data is uniquely suited for framing analysis 
as it allows precise separation of where the pitch actually crossed the plate 
from how the catcher influenced the umpire's call.

The dataset was filtered to called pitches only. Swings, fouls, and balls in 
play were excluded since framing only applies to pitches the umpire must judge. 
Pitches with missing plate coordinates, strike zone boundaries, or catcher ID 
were removed. A shadow zone filter was then applied, retaining only pitches 
within one baseball diameter of the strike zone edge, consistent with MLB's 
official shadow zone definition (MLB.com Glossary). The threshold of 0.242 ft 
was derived from the 2026 Official Baseball Rules circumference specification of 
9.000–9.250 inches. Vertical pitch location was normalized (`pz_norm`) relative 
to each batter's individual strike zone to control for batter height differences. 
Finally, a minimum of 300 shadow zone pitches per catcher was applied to ensure 
reliable random effect estimates, reducing the dataset from 111 to 85 catchers 
and from 171,304 to 168,709 pitches. The final model dataset contains 168,709 
shadow zone pitches across 85 catchers.

## Exploratory Data Analysis

The strike rate for pitches in the shadow zone is 34.2%. This confirms that 
borderline pitches are genuinely ambiguous, umpires do not consistently call them
strikes. A heatmap of called strike rate by pitch location reveals a clear gradient
from near-certain strikes at the interior edge of the shadow zone to near-certain 
balls at the exterior edge, confirming that borderline calls are not deterministic
(Figure 1). The interior of the strike zone is excluded by design since pitches 
well inside the zone are not influenced by framing.

Figure 2 displays the distribution of shadow zone pitches per catcher. This
figure is right-skewed with a long right tail of starting catchers. The 
300-pitch minimum cutoff removes 26 catchers with insufficient data, retaining 
85 catchers with reliable sample sizes for modeling. The median catcher saw 
approximately 1,209 shadow zone pitches across the season, while the most active 
framing catcher received 4,572.

Figure 3 shows that called strike rate varies dramatically by count, ranging 
from 57.0% on 3-0 counts to 10.6% on 0-2 counts. This confirms that count 
strongly influences umpire behavior on borderline pitches — in pitcher-friendly 
counts umpires are more likely to reward a well-located pitch with a strike call, 
while in batter-friendly counts the opposite holds. This pattern justifies 
inclusion of `count` as a fixed effect in the model.

Finally, Figure 4 reveals that right-handed batters consistently receive slightly 
higher called strike rates than left-handed batters across nearly every count, 
with the largest gap observed on 3-0 counts. This persistent handedness effect 
justifies including `stand` as a fixed effect to avoid attributing batter 
handedness differences to catcher framing.                                         

# Methods

## Morgan: Pitch Framing via Mixed-Effects Logistic Regression

### Model Description**

- Goal: isolate catcher influence on borderline pitch calls above and beyond 
  pitch location, count, and batter/pitcher context
- Mixed-effects logistic regression with binary outcome: called strike (1) or ball (0)
- Fixed effects: `plate_x`, `pz_norm`, `count`, `stand`, `p_throws`
- Random effects: `catcher_id`, `pitcher_id`, `batter_id`
- Catcher random effect is the primary quantity of interest — represents each 
  catcher's tendency to influence umpire calls above or below average

### Model Assumptions

- Binary outcome follows a Bernoulli distribution: $Y_i \sim \text{Bernoulli}(p_i)$
- Log-odds of a called strike modeled as a linear function of fixed and random effects
- Catcher, pitcher, and batter random effects are assumed independently normally 
  distributed with mean zero: $u_j \sim N(0, \sigma^2)$
- Random effects are independent of each other and of the fixed effect predictors
- Shadow zone pitches are conditionally independent given the random effects

### Justification

- Mixed-effects over fixed effects: handles the large number of catchers (85), 
  borrows strength across small samples for backup catchers via partial pooling, 
  and shrinks unreliable estimates toward the mean
- Pitcher and batter random effects included to avoid attributing pitcher/batter 
  tendencies to the catcher — directly addresses confounding identified in the 
  assignment description
- BOBYQA optimizer specified due to large dataset size and three-way random effect 
  structure, which can cause premature convergence with the default optimizer

### Model Comparison/Evaluation

- Reduced model fit without pitcher and batter random effects:
  `is_strike ~ plate_x + pz_norm + count + stand + p_throws + (1 | catcher_id)`
- Full vs. reduced model compared via AIC — lower AIC indicates the added 
  complexity of pitcher and batter random effects is justified
- Likelihood ratio test used to formally assess whether adding pitcher and batter 
  random effects significantly improves model fit
- If full model fits better, confirms that omitting pitcher/batter effects would 
  bias catcher framing estimates

### Uncertainty Quantification

- Structured bootstrap resampling following the approach of the course demo
- Unit of resampling: plate appearances within games — replays each game while 
  preserving the schedule structure and pitch dependencies within PAs
- 100 bootstrap iterations, each refitting the full mixed-effects model
- 90% confidence intervals reported for each catcher's random effect estimate
- `set.seed(460)` set before bootstrap loop for reproducibility
- Justification: case bootstrap makes fewer assumptions than parametric bootstrap — 
  does not assume the model is perfectly specified, and resampling at the PA-within-game 
  level reflects the natural source of variability in catcher assignments across a season

# Results

## Morgan: Framing Model Results

- Full model AIC (204,945) is 296 units lower than reduced model AIC (205,241)
- Likelihood ratio test: χ²(2) = 300.25, p < 2.2e-16
- Adding pitcher and batter random effects significantly improves model fit
- Confirms that pitcher and batter tendencies independently influence borderline 
  pitch calls — omitting them would bias catcher framing estimates
- Full model selected for all subsequent inference on catcher framing effects
- Pitcher handedness (`p_throws`) was not significant (p = 0.468), suggesting 
  that after controlling for pitch location and count, pitcher handedness does 
  not meaningfully influence borderline call rates — however it was retained 
  in the model for completeness
  - Catcher variance (0.006) is smaller than pitcher (0.013) and batter (0.016) 
  variance, suggesting location and batter/pitcher context explain more variation 
  in borderline calls than catcher framing alone — but catcher variance is 
  meaningfully non-zero, justifying the framing analysis

# Discussion

## Conclusions

## Limitations
- Shadow zone threshold of 0.242 ft is derived from average baseball diameter — 
  results may be sensitive to this choice, a robustness check with alternative 
  thresholds would strengthen conclusions
- Bootstrap with 100 iterations may not fully capture uncertainty for catchers with 
  smaller sample sizes — increasing to 500-1000 iterations would produce more stable CIs
- Model assumes catcher random effects are normally distributed — this may not hold 
  if framing ability is bimodal (elite framers vs. everyone else)
- Pitcher running game tendencies and wild pitch rates not accounted for — some 
  pitcher effects may still bleed into catcher estimates
- Selection bias not addressed — opposing teams may attempt fewer stolen bases 
  against strong catchers, affecting game context in ways not captured by the model
- Single season analysis — framing estimates based on one season may reflect some 
  noise, year-to-year stability not verified due to scope of project

## Future Work
- Year-to-year stability check using 2024 data — high correlation between 2024 and 
  2025 catcher rankings would confirm the model is measuring real skill vs. noise
- Increase bootstrap iterations to 500-1000 for more stable uncertainty estimates
- Bayesian implementation using `rstanarm` would provide full posterior distributions 
  on catcher effects rather than bootstrap approximations

# References

MLB.com. "Catcher Framing." *MLB Glossary*, Major League Baseball, 
https://www.mlb.com/glossary/statcast/catcher-framing. Accessed April 2026.

MLB Advanced Media. "Catcher Framing Leaderboard." *Baseball Savant*, 
Major League Baseball, 2025. 
https://baseballsavant.mlb.com/catcher_framing. Accessed April 2026.

MLB Advanced Media. *Statcast Pitch-Level Data, 2025 MLB Regular Season.* 
Baseball Savant, Major League Baseball, 2025. 
https://baseballsavant.mlb.com/statcast_search. Accessed April 2026.

Major League Baseball. *2026 Official Baseball Rules*, pg. 5. 
https://mktg.mlbstatic.com/mlb/official-information/2026-official-baseball-rules.pdf. 
Accessed April 2026.

Powell, M.J.D. "The BOBYQA Algorithm for Bound Constrained Optimization Without 
Derivatives." *Cambridge NA Report NA2009/06*, University of Cambridge, 2009.

# Appendix

## Tables

```{r}
#| echo: false
model_comparison_table
```

```{r}
#| echo: false
variance_table
```

## Figures

```{r}
#| echo: false
#| warning: false
#| fig-cap: "Called strike rate by pitch location (2025 shadow zone). Solid box = strike zone, dashed box = shadow zone boundary. Interior of strike zone excluded by design."
#| label: fig-heatmap
#| fig-width: 6
#| fig-height: 5

heatmap_plot
```

```{r}
#| echo: false
#| warning: false
#| fig-cap: "Distribution of shadow zone pitches per catcher (2025 season). Dashed line indicates 300-pitch minimum threshold."
#| label: fig-pitchdist
#| fig-width: 6
#| fig-height: 4

pitchdist_plot
```

```{r}
#| echo: false
#| warning: false
#| fig-cap: "Called strike rate by count in shadow zone (2025 season)."
#| label: fig-countrate
#| fig-width: 6
#| fig-height: 5

countrate_plot
```

```{r}
#| echo: false
#| warning: false
#| fig-cap: "Called strike rate by count and batter handedness (2025 shadow zone)."
#| label: fig-handedness
#| fig-width: 6
#| fig-height: 5

handedness_plot
```

