---
title: "Draft: Measuring MLB Catcher Defensive Contributions"
author: "Arthur Spirou, Joel Maison, and Morgan Corliss"
date: "15 April 2026"
format: 
  pdf:
    toc: false
execute:
  echo: false
  warning: false
  message: false
editor: 
  markdown: 
    wrap: 72
---

```{r}
#| echo: false
#| warning: false
#| message: false

library(tidyverse)
library(knitr)

# Load all saved objects
heatmap_plot            <- readRDS("heatmap_plot.rds")
pitchdist_plot          <- readRDS("pitchdist_plot.rds")
countrate_plot          <- readRDS("countrate_plot.rds")
handedness_plot         <- readRDS("handedness_plot.rds")
caterpillar_plot        <- readRDS("caterpillar_plot.rds")
runssaved_plot          <- readRDS("runssaved_plot.rds")
model_comparison_table  <- readRDS("model_comparison_table.rds")
variance_fixed_table    <- readRDS("variance_fixed_table.rds")
leaderboard_table       <- readRDS("leaderboard_table.rds")
```

# Abstract

Catcher defense is among the most difficult contributions to measure in baseball. Traditional statistics such as passed balls and caught stealing rate capture only isolated aspects of a catcher's defensive value, leaving a large portion of their impact unquantified. We develop a hierarchical modeling framework that isolates each catcher's defensive contribution across three dimensions: pitch framing, pitch blocking, and count-specific framing effects. For framing, we fit a mixed-effects logistic regression model to 168,709 shadow zone pitches from the 2025 MLB regular season, extracting catcher random effects that represent each catcher's influence on borderline pitch calls after controlling for pitch location, count, and pitcher and batter tendencies. For blocking, we apply a similar model to 157,195 below-zone pitches, isolating each catcher's ability to convert difficult pitches into recorded blocks. We further examine whether catcher framing ability varies by count, finding statistically significant evidence that some catchers are especially effective at influencing borderline calls in specific game situations. Model comparisons via AIC and likelihood ratio tests confirm that catcher identity explains meaningful variation in both framing and blocking outcomes above and beyond contextual controls.

# Introduction

In baseball, catchers play a pivotal role on defense, yet their value is
historically difficult to measure. Traditional statistics such as passed
balls and caught stealing rate capture only a fraction of what a catcher
contributes. In this project we focus on three key aspects of catcher
defense: pitch framing, pitch blocking, and count-specific framing
effects. We then quantified each in terms of runs saved, an easy to
understand metric that allows for direct comparison across defensive
skills and is accessible to even the most casual baseball fan.

Pitch framing, the ability to influence umpires to call borderline
pitches strikes, is one of the most impactful and measurable aspects of
catcher defense. A single extra strike per game can shift run
expectancy, heighten pitcher confidence, and alter game outcomes. Elite
framers can provide significant value, in the 2024 MLB season, the
top-rated framing catcher Patrick Bailey saved 22 runs through framing
alone, with the next five catchers saving between 11 and 13 runs (MLB
Advanced Media, 2025). With the rise of pitch tracking, pitch
coordinates are now widely available, framing is another way teams can
evaluate catchers to ensure their getting the best guys for the job.

To measure framing, we fit a mixed-effects logistic regression model to
168,709 shadow zone pitches from the 2025 MLB season, isolating each
catcher's framing contribution as a random effect after controlling for
pitch location, count, batter, and pitcher. Results reveal meaningful
variation in framing ability across MLB catchers. \[Pending bootstrap
completion — uncertainty estimates and catcher rankings to be added
prior to final submission.

Blocking potentially wild pitches, swinging strikes, and balls that are
low and/or off the plate is an extremely important defensive trait for
catchers. This is especially true when runners are on base, because
successfully blocking a ball can prevent immediate runs or runner
advancement on a given pitch while also shaping the rest of the inning
by keeping runners from moving into more advantageous positions.
Traditionally, Baseball Savant has used advanced inputs such as catcher
positioning and called pitch type, which are not available in public
pitch-level data, to build catcher blocking metrics and rankings. We aim
to develop a comparable model for ranking catcher blocking ability using
only publicly available pitch-level data.

Blocking pitches has effects that go beyond the box score. In addition
to directly preventing passed balls, wild pitches, runs, and stolen
bases, strong blocking can also influence pitcher and team confidence.
When a pitcher trusts their catcher to handle balls in the dirt or
pitches that break well outside the strike zone, they are better able to
use their full arsenal aggressively against hitters. Many of the most
effective pitches are designed to look hittable before finishing in
locations that are difficult or impossible to square up, often below the
zone or well off the plate. In high-leverage situations against strong
hitters, the ability to throw those pitches with confidence can have a
major impact on game outcomes and on a pitcher’s overall effectiveness.
Because many catchers are valued more for their defense than their
offense, methods for evaluating pitch-blocking skill are especially
important for teams. Quantifying blocking ability can improve catcher
evaluation on its own, while also supporting better analysis of pitchers
and providing teams with additional insight into the effectiveness of
specific pitcher-catcher battery combinations.

Across all three analyses, our models confirm that catcher identity
explains meaningful variation in defensive outcomes above and beyond
contextual controls. The framing model reveals significant variation in
catcher ability to influence borderline pitch calls, with non-zero
catcher variance confirmed after controlling for pitch location, count,
pitcher, and batter effects. The blocking model similarly finds that
catcher identity explains variation in recorded block rates above and
beyond pitch type and location, though pitcher tendencies account for
more variation than catcher differences alone. Finally, count-specific
framing effects are statistically significant, suggesting that some
catchers are especially effective at converting borderline pitches into
called strikes in specific count situations. Full uncertainty estimates
and catcher rankings are pending bootstrap completion prior to final
submission.

# Data

Pitch-level Statcast data was accessed via Baseball Savant's CSV export
API for the 2025 MLB regular season (March 27 – September 28, 2025),
yielding 712,192 total pitches across 118 variables (MLB Advanced Media,
2025). Key variables include `plate_x` and `plate_z` (pitch coordinates
at home plate), `sz_top` and `sz_bot` (batter-specific strike zone
boundaries), `fielder_2` (catcher ID), `description` (called strike or
ball), `balls`, `strikes`, `stand`, and `p_throws`. Pitch-level
coordinate data is uniquely suited for framing analysis as it allows
precise separation of where the pitch actually crossed the plate from
how the catcher influenced the umpire's call.

Key variables for the blocking analysis: plate_x, plate_z (pitch
coordinates), sz_top, sz_bot (batter-specific strike zone boundaries),
fielder_2 (catcher ID), pitch_type, release_speed, description
(including blocked_ball and swinging_strike_blocked), balls, strikes,
p_throws, and des (used to identify official wild pitch / passed ball
outcomes for a secondary check)

## Preprocessing: Catcher Framing

The dataset was filtered to called pitches only. Swings, fouls, and
balls in play were excluded since framing only applies to pitches the
umpire must judge. Pitches with missing plate coordinates, strike zone
boundaries, or catcher ID were removed. A shadow zone filter was then
applied, retaining only pitches within one baseball diameter of the
strike zone edge, consistent with MLB's official shadow zone definition
(MLB.com Glossary). The threshold of 0.242 ft was derived from the 2026
Official Baseball Rules circumference specification of 9.000–9.250
inches. Vertical pitch location was normalized (`pz_norm`) relative to
each batter's individual strike zone to control for batter height
differences. Finally, a minimum of 300 shadow zone pitches per catcher
was applied to ensure reliable random effect estimates, reducing the
dataset from 111 to 85 catchers and from 171,304 to 168,709 pitches. The
final model dataset contains 168,709 shadow zone pitches across 85
catchers.

## Preprocessing: Catcher Blocking

We filtered the data to pitches with non-missing values for plate_x,
plate_z, sz_top, sz_bot, fielder_2, pitch_type, and release_speed. We
then defined blocking opportunities as pitches below the strike zone,
where plate_z \< sz_bot, to isolate pitches on which a block was
plausibly required. A binary outcome variable, is_block, was created and
set equal to 1 when the pitch description was blocked_ball or
swinging_strike_blocked, and 0 otherwise, allowing us to define block
rate as is_block over blocking opportunities. We also renamed fielder_2
and pitcher to catcher_id and pitcher_id, created a count variable from
the balls-strikes combination, and formatted pitch_type and p_throws as
categorical variables. To avoid unstable coefficient estimates, pitch
types with fewer than 500 below-zone pitches were grouped into an
"Other" category. We considered including a swing indicator, but removed
it because swinging_strike_blocked already implies a recorded block. The
resulting simple blocking model dataset contains 157,195 below-zone
pitches across 96 catchers and 857 pitchers. Finally, we used a
catcher-inclusion threshold of 100 below-zone pitches, since blocking
opportunities are much rarer than framing opportunities.

## Exploratory Data Analysis: Catcher Framing

The strike rate for pitches in the shadow zone is 34.2%. This confirms
that borderline pitches are genuinely ambiguous, umpires do not
consistently call them strikes. A heatmap of called strike rate by pitch
location reveals a clear gradient from near-certain strikes at the
interior edge of the shadow zone to near-certain balls at the exterior
edge, confirming that borderline calls are not deterministic
(@fig-heatmap). The interior of the strike zone is excluded by design
since pitches well inside the zone are not influenced by framing.

@fig-pitchdist displays the distribution of shadow zone pitches per
catcher. This figure is right-skewed with a long right tail of starting
catchers. The 300-pitch minimum cutoff removes 26 catchers with
insufficient data, retaining 85 catchers with reliable sample sizes for
modeling. Catchers retained for modeling saw anywhere from 300 to
approximately 4,500 shadow zone pitches.

@fig-countrate shows that called strike rate varies dramatically by
count, ranging from 57% on 3-0 counts to 10.6% on 0-2 counts. This
confirms that count strongly influences umpire behavior on borderline
pitches. Notably, the highest strike rates occur on hitter-friendly
counts like 3-0 and 2-0, where umpires appear more willing to call
borderline pitches strikes when the pitcher needs to throw one, while
two-strike counts like 0-2 and 1-2 show the lowest strike rates. This
pattern justifies inclusion of `count` as a fixed effect in the model to
avoid attributing count effects to catcher framing.

Finally, @fig-handedness reveals that right-handed batters consistently
receive higher called strike rates than left-handed batters across every
count with the largest gap observed on 3-1 counts. This persistent
handedness effect justifies including `stand` as a fixed effect to avoid
attributing batter handedness differences to catcher framing.

## Exploratory Data Analysis: Catcher Blocking

Inspection of the description field confirmed that Statcast includes
explicit labels for successful blocks through blocked_ball and
swinging_strike_blocked. By contrast, official blocking failures appear
extremely sparse in publicly available pitch-level data: searching the
des field identified only 86 wild pitches and 29 passed balls. This
strong imbalance suggests that official wild pitch and passed ball
outcomes substantially undercount true blocking failures, since many
missed or difficult blocks are not recorded as official failures. For
example, a passed ball with no runners on base may simply be scored as a
ball rather than as a passed ball. Because of this limitation, recorded
block rate among below-zone opportunities will serve as the primary
modeling outcome, while official wild pitch and passed ball outcomes
will be used only in a secondary supplemental model for evaluation.
Blocking opportunities are also distributed unevenly across catchers,
with a long tail of low-volume catchers, so a 100-pitch cutoff helps
stabilize catcher-level random-effect estimates. In addition, there is
substantial variation in below-zone opportunity counts across pitch
types, which is why sparse pitch types are collapsed into an Other
category before model fitting.

# Methods

## Morgan Corliss: Catcher Framing

The goal of this analysis is to isolate each catcher's influence on
borderline pitch calls above and beyond what can be explained by pitch
location, count, and batter/pitcher context. To accomplish this, we fit
a mixed-effects logistic regression model where the binary outcome is
whether a shadow zone pitch was called a strike (1) or a ball (0). The
model takes the following form:

$$\log\left(\frac{p_i}{1-p_i}\right) = \beta_0 + \beta_1 \text{plate\_x}_i + 
\beta_2 \text{pz\_norm}_i + \beta_3 \text{count}_i + \beta_4 \text{stand}_i + 
\beta_5 \text{p\_throws}_i + u_{\text{catcher}[i]} + u_{\text{pitcher}[i]} + 
u_{\text{batter}[i]}$$

where $p_i$ is the probability of a called strike on pitch $i$. The
fixed effects control for pitch location (`plate_x`, `pz_norm`), count,
batter handedness (`stand`), and pitcher handedness (`p_throws`). The
random effects $u_{\text{catcher}[i]}$, $u_{\text{pitcher}[i]}$, and
$u_{\text{batter}[i]}$ capture the influence of each catcher, pitcher,
and batter respectively. The catcher random effect is the primary
quantity of interest — it represents each catcher's tendency to
influence umpire calls above or below the league average after
accounting for all other factors.

The model assumes the binary outcome follows a Bernoulli distribution:
$Y_i \sim \text{Bernoulli}(p_i)$, where the log-odds of a called strike
are modeled as a linear function of the fixed and random effects. The
catcher, pitcher, and batter random effects are each assumed to be
independently normally distributed with mean zero:
$u_j \sim N(0, \sigma^2)$. Random effects are assumed independent of
each other and of the fixed effect predictors, and shadow zone pitches
are assumed conditionally independent given the random effects.

A mixed-effects model was chosen over a fixed-effects specification for
several reasons. With 85 catchers in the dataset, a fixed-effects
approach would require estimating 84 catcher indicator coefficients,
producing unreliable estimates for backup catchers with small sample
sizes. The mixed-effects framework handles this naturally through
partial pooling — shrinking estimates for catchers with limited data
toward the league mean while allowing well-sampled catchers to deviate
further. Pitcher and batter random effects were included to avoid
attributing their tendencies to catchers, which would otherwise bias the
framing estimates. The BOBYQA optimizer was specified following
recommendations for mixed models with complex random effect structures
(Powell, 2009).

To evaluate the model, we fit a reduced specification excluding the
pitcher and batter random effects:

$$\text{is\_strike} \sim \text{plate\_x} + \text{pz\_norm} + \text{count} + 
\text{stand} + \text{p\_throws} + (1 | \text{catcher\_id})$$

The full and reduced models were compared via AIC, where a lower value
indicates better fit relative to model complexity. A likelihood ratio
test was also conducted to formally assess whether adding pitcher and
batter random effects significantly improves model fit. If the full
model fits better, it confirms that omitting pitcher and batter effects
would bias the catcher framing estimates.

Uncertainty in catcher random effect estimates is quantified using conditional 
standard errors from the mixed-effects model, following the approach demonstrated 
in course materials. Ninety percent confidence intervals are computed 
as the estimate ± 1.645 standard errors. These intervals reflect uncertainty 
under the assumption that the model is correctly specified.

## Arthur Spirou: Catcher Blocking via Mixed-Effects Logistic Regression

### Blocking Model Description\*\*

-   Goal: isolate catcher influence on likely blocking opportunities
    above and beyond pitch location, pitch type, and pitcher context
-   Mixed-effects logistic regression with binary outcome: recorded
    block (1) or no recorded block (0)
-   Primary outcome defined as is_block = 1 when description was
    blocked_ball or swinging_strike_blocked
-   Restricted to below-zone pitches (plate_z \< sz_bot)
-   Simple model fixed effects: plate_x, plate_z, and collapsed pitch
    type
-   Simple model random effects: catcher_id, pitcher_id
-   Full model will build on the simple model to include release_speed,
    count, and p_throws
-   Catcher random effect represents each catcher’s ability to convert
    likely blocking opportunities into recorded blocks, relative to
    league average. Controlling for block attempt difficulty and pitcher

### Blocking Model Assumptions

-   Binary outcome follows a Bernoulli distribution
-   log-odds of a recorded block modeled as a linear function of fixed
    and random effects
-   catcher and pitcher random effects are assumed to be independent.
    Also assumed to be normally distributed with mean zero
-   random effects are independent of fixed effects
-   The below-zone filter is assumed to capture a meaningful set of
    likely blocking opportunities, however it does not perfectly
    identify every true blocking situation
-   Outcome is interpreted as a recorded block rate, not as a fully
    observed success-versus-failure blocking label

Simple Blocking Model Formulation:

$$
Y_i \mid u_{j(i)}, v_{k(i)} \sim \mathrm{Bernoulli}(p_i)
$$

$$
\log\left(\frac{p_i}{1 - p_i}\right)
=
\beta_0
+ \beta_1 \,\mathrm{plate\_x}_i
+ \beta_2 \,\mathrm{plate\_z}_i
+ \boldsymbol{\gamma}^{\top}\mathbf{z}_i
+ u_{j(i)}
+ v_{k(i)}
$$

$$
u_j \sim N(0, \sigma^2_{\mathrm{catcher}})
$$

$$
v_k \sim N(0, \sigma^2_{\mathrm{pitcher}})
$$

Here, $Y_i$ indicates whether below-zone pitch $i$ resulted in a
recorded block, $\mathbf{z}_i$ is the vector of indicator variables for
`pitch_type_collapsed`, $u_{j(i)}$ is the catcher-specific random
intercept, and $v_{k(i)}$ is the pitcher-specific random intercept.

### Blocking Model Justification

-   Mixed-effects logistic regression is appropriate because the outcome
    is binary. Also, repeated observations from the same catchers and
    pitchers create dependence
-   Treating catcher as a random effect allows partial pooling, which
    stabilizes estimates for lower-volume catchers and avoids
    overfitting
-   Pitcher randomeffects are particularly crucial and meaningful for a
    blocking model because some pitchers throw balls in the dirt or
    balls that move dramatically away from the strike zone more often
    than others. This can both create more blocking opportunities and
    pose more challenging blocking opportunities for certain catchers.
-   Similar to pitch framing model, BOBYQA optimizer specified due to
    large dataset size and crossed random effects nature

### Blocking Model Comparison/Evaluation

Simple Model: is_block \~ plate_x + plate_z + pitch_type_collapsed + (1
\| catcher_id) + (1 \| pitcher_id) Full Candidate Model: is_block \~
plate_x + plate_z + pitch_type_collapsed + release_speed + count +
p_throws + (1 \| catcher_id) + (1 \| pitcher_id) - Plan to compare the
simple and full models using AIC, where lower AIC indicates better
out-of-sample tradeoff between fit and complexity - A likelihood ratio
test will be used to assess whether the additional fixed effects in the
full model significantly improve fit - Catcher effects will be ranked
and compared to Baseball Savant's catcher blocking ratings from the 2025
season

### Blocking Model Uncertainty Quantification

-   Similar to the framing model, we plan to use structured bootstrap
    resampling with the unit of resampling being plate appearances
    within games to preserve within-game depe ndence and catcher usage
    patterns
-   Each bootstrap sample will refit the selected blocking model and
    store the catcher random effects
-   Plan to report 90% confidence intervals for catcher blocking
    effects. Random seed will be used for reproducibility.

## Joel Maison: Count-Specific Catcher Framing via Mixed-Effects Logistic Regression

Prior work on catcher framing typically estimates a single defensive value per catcher, treating framing ability as
constant across all game situations. This analysis asks a more granular
question: does a catcher's ability to influence borderline pitch calls
vary depending on the count? A 3-0 count and a 0-2 count create very
different contexts for umpire decision making, and elite framers may be
especially effective in high leverage situations where stealing a strike
matters most.

To test this, I fit a mixed-effects logistic regression model on 168,709
shadow zone pitches from the 2025 MLB regular season with a binary
outcome of called strike (1) or ball (0). Fixed effects control for
pitch location (plate_x, pz_norm), count, batter handedness (stand), and
pitcher handedness (p_throws). The model includes catcher, pitcher, and
batter random intercepts, and adds a catcher-by-count interaction random
effect as the primary quantity of interest:

$$\log\left(\frac{p_i}{1-p_i}\right) = \beta_0 + \beta_1 \text{plate\_x}_i + 
\beta_2 \text{pz\_norm}_i + \beta_3 \text{count}_i + \beta_4 \text{stand}_i + 
\beta_5 \text{p\_throws}_i + u_{\text{catcher}[i]} + 
u_{\text{catcher}[i]\times\text{count}[i]} + u_{\text{pitcher}[i]} + 
u_{\text{batter}[i]}$$

The term $u_{\text{catcher}[i]\times\text{count}[i]}$ captures each
catcher's deviation from their own average framing ability within each
specific count, and is the primary quantity of interest. The remaining
random effects $u_{\text{catcher}[i]}$, $u_{\text{pitcher}[i]}$, and
$u_{\text{batter}[i]}$ control for overall catcher, pitcher, and batter
tendencies respectively.

The model assumes the binary outcome follows a Bernoulli distribution:
$Y_i \sim \text{Bernoulli}(p_i)$, where the log-odds of a called strike
are modeled as a linear function of the fixed and random effects. All
random effects are assumed to be independently normally distributed with
mean zero: $u_j \sim N(0, \sigma^2)$. Random effects are assumed
independent of each other and of the fixed effect predictors, and shadow
zone pitches are assumed conditionally independent given the random
effects.

A mixed-effects specification was chosen for the same reasons as the
framing model: partial pooling stabilizes estimates for catchers with
limited data by shrinking them toward the league mean, while allowing
well-sampled catchers to deviate further. The catcher-by-count
interaction is modeled as a random effect rather than a fixed effect
because estimating a fixed coefficient for every catcher-count
combination would require 1,020 parameters, producing highly unreliable
estimates, particularly for backup catchers who face few pitches in rare
counts. The random effect framework pools information across all
catcher-count cells, producing stable estimates even for sparse
combinations. The BOBYQA optimizer was specified following
recommendations for mixed models with complex random effect structures
(Powell, 2009).

To evaluate whether the catcher-by-count interaction meaningfully
improves model fit, we compare our model against a baseline
specification without the interaction term, equivalent to the baseline
framing model:

$$\text{is\_strike} \sim \text{plate\_x} + \text{pz\_norm} + \text{count} + 
\text{stand} + \text{p\_throws} + (1 | \text{catcher\_id}) + 
(1 | \text{pitcher\_id}) + (1 | \text{batter\_id})$$

The two models are compared via AIC, where a lower value indicates
better fit relative to model complexity. A likelihood ratio test is also
conducted to formally assess whether the interaction term significantly
improves fit. If the interaction model fits better, it confirms that
catcher framing ability varies meaningfully across counts above and
beyond overall framing skill.

To quantify uncertainty in the catcher-by-count random effect estimates,
we will employ the same structured bootstrap resampling approach used in
the framing model, resampling plate appearances within games to preserve
within-game dependence. Each bootstrap iteration refits the full model
and stores the catcher-by-count random effects. We will report 90%
bootstrap confidence intervals for each catcher-count combination, with
set.seed(460) for reproducibility.

# Results

## Morgan Corliss: Catcher Framing

### Model Comparison

The full model achieved a substantially lower AIC (204,945) compared to
the reduced model (205,241), a difference of 296 units — well above the
conventional threshold of 2. The likelihood ratio test confirms this
improvement is statistically significant (χ²(2) = 300.25, p \< 2.2e-16),
providing strong evidence that pitcher and batter random effects
meaningfully improve model fit. This result confirms that pitcher and
batter tendencies independently influence borderline pitch calls above
and beyond pitch location and count — omitting them would attribute some
of their influence to catchers, biasing the framing estimates. The full
model was therefore selected for all subsequent inference on catcher
framing effects. These results are summarized in
(@tbl-model-comparison).

### Model Interpretation

@tbl-variance-fixed displays the variance components and key fixed
effects from the full model. Catcher variance (0.006) is smaller than
both pitcher (0.013) and batter (0.016) variance, suggesting that pitch
location and batter/pitcher context explain more variation in borderline
calls than catcher framing alone. However, catcher variance is
meaningfully non-zero, confirming that framing ability varies across
catchers above and beyond these controls and justifying the framing
analysis.

Among the fixed effects, pitcher handedness (`p_throwsR`) was not
statistically significant (β = -0.012, p = 0.468), suggesting that after
controlling for pitch location and count, pitcher handedness does not
meaningfully influence borderline call rates. It was retained in the
model for completeness. All count coefficients were highly significant
(p \< 2e-16), consistent with the dramatic variation in called strike
rates across counts observed in the EDA — with 3-0 counts showing the
highest tendency toward called strikes (β = 0.586) and 0-2 counts the
lowest (β = -1.897). Right-handed batters received significantly more
called strikes on borderline pitches than left-handed batters (β =
0.140, p \< 2e-16), consistent with the handedness effect observed in
(@fig-handedness).

### Catcher Framing Estimates

\[Pending bootstrap completion — caterpillar plot, ridge plot, runs
saved leaderboard, and uncertainty estimates to be added prior to final
submission\]

## Arthur Spirou: Catcher Blocking Results

The simple blocking model converged successfully on 157,195 below-zone
pitches drawn from 96 catchers and 857 pitchers. Both catcher- and
pitcher-level random-effect variances were significant, indicating that
blocking outcomes vary across both groups even after controlling for
pitch type and pitch location. Notably, the pitcher-level variance
(0.059) exceeded the catcher-level variance (0.022), suggesting that
pitcher tendencies and overall pitch difficulty explain more variation
in recorded blocks than catcher differences alone. In the simple model,
vertical pitch location (plate_z) was the dominant predictor of a
recorded block (estimate = -2.218, p \< 2e-16), meaning that lower
pitches were much more likely to result in a block. Horizontal pitch
location (plate_x) was also a significant predictor (estimate = -0.061,
p \< 0.001). In addition, different types of breaking balls were
associated with higher odds of a recorded block, while four-seam
fastballs and sinkers had the lowest odds. The full model will be
compared with the simple model using AIC and a likelihood ratio test.
Finally, official wild pitch and passed ball labels were found to be
extremely sparse, with only 86 wild pitches and 29 passed balls
identified, further reinforcing the decision not to use official
failures as the primary blocking outcome.

## Joel Maison: Count-Specific Catcher Framing

### Model Comparison

The interaction model achieved a lower AIC (204,937) compared to the
baseline model without the catcher-by-count interaction (204,945), a
difference of 8 units. The likelihood ratio test confirms this
improvement is statistically significant (χ²(1) = 9.457, p = 0.002),
providing evidence that adding the catcher-by-count interaction
meaningfully improves model fit above and beyond overall framing skill.
This result confirms that catcher framing ability is not constant across
all counts and that the game situation influences how effectively
individual catchers convert borderline pitches into called strikes.
These results are summarized in (@tbl-count-comparison).

### Model Interpretation

The catcher-by-count interaction variance (0.0038) is smaller than the
overall catcher variance (0.0056), suggesting that while catchers do
vary in their framing ability across counts, a larger portion of
catcher-level variation is captured by their overall framing skill.
Nevertheless, the non-zero interaction variance combined with the
significant likelihood ratio test confirms that count-specific framing
effects exist and are not simply noise.

The fixed effects are consistent with the EDA findings. Called strike
rates drop dramatically from 3-0 counts (β = 0.584, p \< 2e-16) to 0-2
counts (β = -1.900, p \< 2e-16), confirming that count strongly shapes
umpire behavior on borderline pitches. Right-handed batters received
significantly more called strikes than left-handed batters (β = 0.140, p
\< 2e-16). Pitcher handedness was not significant (β = -0.012, p =
0.472), consistent with Morgan's framing model findings. The top and bottom 
catcher-count combinations are displayed in (@fig-count-effects).

### Uncertainty

\[Pending bootstrap completion — confidence intervals for
catcher-by-count random effects to be added prior to final submission.\]

# Discussion

## Conclusions

The model comparison results provide strong evidence that catcher
framing is a real and measurable skill. The full mixed-effects model fit
significantly better than the reduced model (AIC difference = 296, χ²(2)
= 300.25, p \< 2.2e-16), confirming that catcher identity explains
meaningful variation in borderline pitch calls above and beyond pitch
location, count, pitcher, and batter effects. The non-zero catcher
variance (0.006) further supports this conclusion.

Count-specific framing effects were also found to be statistically
significant (χ²(1) = 9.457, p = 0.002), suggesting that catcher framing
ability is not uniform across all game situations and that some catchers
may be especially effective in high leverage counts.

*\[Pending bootstrap completion — conclusions about the magnitude of
framing differences across catchers and runs saved estimates to be added
prior to final submission.\]*

## Limitations

Several limitations of this analysis should be noted. First, the shadow
zone threshold of 0.242 ft is derived from the average baseball diameter
per the 2026 Official Baseball Rules and a robustness check with
alternative values would strengthen this analysis. Second, this analysis
is based on a single season of data, framing estimates from one season
may reflect some noise. Conducting an analysis of different years would
help verify our conclusions.

The blocking analysis has several important limitations. First, the
blocking outcome is not a perfect success-versus-failure label: the 0
class includes below-zone pitches that did not receive a recorded block,
but some of those pitches may have been received cleanly rather than
missed. Official wild pitch and passed ball outcomes are too sparse to
serve as the primary response variable, and they likely undercount true
blocking failures because many difficult or mishandled pitches are never
scored as official failures. In addition, the below-zone filter (plate_z
\< sz_bot) is only a practical approximation of blocking opportunities.
While it helps isolate pitches where a block is plausibly required, it
does not account for pitches far above the zone, may exclude some
horizontally difficult pitches, and may still include low pitches that
were relatively easy for the catcher to receive. The model is also
limited to publicly available pitch-level variables and therefore cannot
incorporate features such as catcher starting position, body
orientation, glove-side versus backhand mechanics, or other tracking
information used in proprietary blocking models like Baseball Savant's.
Finally, pitch type was collapsed for sparse categories to improve model
stability, but doing so may hide meaningful differences among rare pitch
types, such as an eephus pitch.

The count-specific framing analysis is limited by the relatively small
catcher-by-count interaction variance (0.0038), which while
statistically significant is modest in magnitude. Bootstrap confidence
intervals are also pending, so the practical significance of individual
catcher-count differences has not yet been fully quantified.

## Future Work

Several extensions of this work would strengthen the conclusions. A
year-to-year stability check using 2024 data would help determine
whether catcher rankings are consistent across seasons. High correlation
between 2024 and 2025 estimates would provide evidence that the model is
measuring real framing skill rather than noise. Finally, a Bayesian
implementation using `rstanarm` would provide full posterior
distributions on catcher framing effects rather than bootstrap
approximations, accounting for the hierarchical structure of the data.
Future work could also examine which specific counts show the largest
variation in catcher framing effects, and whether count-specific framing
ability is stable year to year.

# References

MLB.com. "Catcher Framing." *MLB Glossary*, Major League Baseball,
https://www.mlb.com/glossary/statcast/catcher-framing. Accessed April
2026.

MLB Advanced Media. "Catcher Framing Leaderboard." *Baseball Savant*,
Major League Baseball, 2025.
https://baseballsavant.mlb.com/catcher_framing. Accessed April 2026.

MLB Advanced Media. *Statcast Pitch-Level Data, 2025 MLB Regular
Season.* Baseball Savant, Major League Baseball, 2025.
https://baseballsavant.mlb.com/statcast_search. Accessed April 2026.

Major League Baseball. *2026 Official Baseball Rules*, pg. 5.
https://mktg.mlbstatic.com/mlb/official-information/2026-official-baseball-rules.pdf.
Accessed April 2026.

Powell, M.J.D. "The BOBYQA Algorithm for Bound Constrained Optimization
Without Derivatives." *Cambridge NA Report NA2009/06*, University of
Cambridge, 2009.

# Appendix

## Tables

```{r}
#| echo: false
#| label: tbl-model-comparison
#| tbl-cap: "Table 1: Model comparison — full vs. reduced mixed-effects logistic regression"

model_comparison_table
```

```{r}
#| echo: false
#| label: tbl-variance-fixed
#| tbl-cap: "Table 2: Variance components and key fixed effects from full mixed-effects model"

variance_fixed_table

```

```{r}
#| echo: false
#| label: tbl-count-comparison
#| tbl-cap: "Table 3: Model comparison — baseline framing model vs. count-specific interaction model"

tibble(
  Model            = c("Baseline model", "Count interaction model"),
  `Random effects` = c(
    "Catcher + Pitcher + Batter",
    "Catcher + Pitcher + Batter + Catcher x Count"
  ),
  AIC              = c(204944.9, 204937.5),
  `Delta AIC`      = c(7.4, 0.0),
  `Chi-sq`         = c("—", "9.457"),
  df               = c("—", "1"),
  `p-value`        = c("—", "0.002")
) %>%
  kable(
    format  = "simple",
    caption = "Model comparison: baseline vs. count-specific interaction model",
    digits  = 1
  )
```

```{r}
#| echo: false
#| label: tbl-leaderboard
#| tbl-cap: "Table 4: Top and bottom 10 catchers by framing runs saved (2025 season)"

leaderboard_table
```

## Figures

```{r}
#| echo: false
#| warning: false
#| fig-cap: "Figure 1: Called strike rate by pitch location (2025 shadow zone). Solid box = strike zone, dashed box = shadow zone boundary. Interior of strike zone excluded by design."
#| label: fig-heatmap
#| fig-width: 6
#| fig-height: 5

heatmap_plot
```

```{r}
#| echo: false
#| warning: false
#| fig-cap: "Figure 2: Distribution of shadow zone pitches per catcher (2025 season). Dashed line indicates 300-pitch minimum threshold."
#| label: fig-pitchdist
#| fig-width: 6
#| fig-height: 4

pitchdist_plot
```

```{r}
#| echo: false
#| warning: false
#| fig-cap: "Figure 3: Called strike rate by count in shadow zone (2025 season)."
#| label: fig-countrate
#| fig-width: 6
#| fig-height: 5

countrate_plot
```

```{r}
#| echo: false
#| warning: false
#| fig-cap: "Figure 4: Called strike rate by count and batter handedness (2025 shadow zone)."
#| label: fig-handedness
#| fig-width: 6
#| fig-height: 5

handedness_plot
```

```{r}
#| echo: false
#| warning: false
#| label: fig-caterpillar
#| fig-cap: "Figure 5: Catcher framing random effects with 90% confidence intervals (2025 season). Red points indicate catchers significantly different from zero."
#| fig-width: 7
#| fig-height: 6

caterpillar_plot
```

```{r}
#| echo: false
#| warning: false
#| label: fig-runssaved
#| fig-cap: "Figure 6: Catcher framing value for top and bottom 10 catchers (2025 season). Error bars show 90% confidence intervals."
#| fig-width: 7
#| fig-height: 6

runssaved_plot
```

```{r}
#| echo: false
#| label: fig-count-effects
#| fig-cap: "Figure 7: Top and bottom 15 catcher-by-count random effects (2025 season). Each point represents a catcher-count combination. Positive values indicate above-average framing in that count."
#| fig-width: 7
#| fig-height: 5

# library(tidyverse)
# library(lme4)
# 
# ranef(framing_count)$`catcher_id:count` %>%
#   rownames_to_column("catcher_count") %>%
#   rename(effect = `(Intercept)`) %>%
#   arrange(desc(effect)) %>%
#   slice(c(1:15, (n()-14):n())) %>%
#   mutate(
#     catcher_count = fct_reorder(catcher_count, effect),
#     direction = ifelse(effect > 0, "Above average", "Below average")
#   ) %>%
#   ggplot(aes(x = effect, y = catcher_count, color = direction)) +
#   geom_point(size = 2.5) +
#   geom_vline(xintercept = 0, linetype = "dashed", linewidth = 0.5) +
#   scale_color_manual(
#     values = c("Above average" = "steelblue", "Below average" = "firebrick"),
#     name = NULL
#   ) +
#   labs(
#     title    = "Catcher-by-count framing effects (2025 season)",
#     subtitle = "Top and bottom 15 catcher-count combinations",
#     x        = "Random effect (log-odds scale)",
#     y        = NULL
#   ) +
#   theme_bw() +
#   theme(
#     panel.grid.minor   = element_blank(),
#     panel.grid.major.y = element_blank(),
#     legend.position    = "bottom"
#   )
```
