---
title: "Measuring MLB Catcher Defensive Contributions"
author: "Arthur Spirou, Joel Maison, and Morgan Corliss"
date: "28 April 2026"
format:
  pdf:
    toc: false
    fontsize: 11pt
    geometry: margin=1in
    fig-pos: 'H'
    include-in-header:
      text: |
        \usepackage{float}
execute:
  echo: false
  warning: false
  message: false
---

\newpage

```{r}
#| echo: false
#| warning: false
#| message: false
#| output: false

library(tidyverse)
library(knitr)

# Load all saved objects
heatmap_plot            <- readRDS("heatmap_plot.rds")
pitchdist_plot          <- readRDS("pitchdist_plot.rds")
countrate_plot          <- readRDS("countrate_plot.rds")
handedness_plot         <- readRDS("handedness_plot.rds")
caterpillar_plot        <- readRDS("caterpillar_plot.rds")
runssaved_plot          <- readRDS("runssaved_plot.rds")
model_comparison_table  <- readRDS("model_comparison_table.rds")
variance_fixed_table    <- readRDS("variance_fixed_table.rds")
leaderboard_table       <- readRDS("leaderboard_table.rds")

# Joel's context dependent framing objects
framing_eda_count_pt_plot      <- readRDS("framing_eda_count_pt_plot.rds")
framing_main_plot              <- readRDS("framing_main_plot.rds")
framing_pt_topbot_plot         <- readRDS("framing_pt_topbot_plot.rds")
framing_pt_profile_plot        <- readRDS("framing_pt_profile_plot.rds")
lrt_summary                    <- readRDS("lrt_summary.rds")
variance_components_table      <- readRDS("variance_components_table.rds")

# Arthur's blocking objects
block_model_comparison_table       <- readRDS("block_model_comparison_table.rds")
block_variance_fixed_table         <- readRDS("block_variance_fixed_table.rds")
catcher_effects_named              <- readRDS("catcher_effects_named.rds")
block_opportunity_comparison_table <- readRDS("block_opportunity_comparison_table.rds")
block_heatmap_plot                 <- readRDS("block_heatmap_plot.rds")
block_pitchdist_plot               <- readRDS("block_pitchdist_plot.rds")
block_by_pitchtype_plot            <- readRDS("block_by_pitchtype_plot.rds")
block_by_depth_plot                <- readRDS("block_by_depth_plot.rds")
block_caterpillar_ci_plot          <- readRDS("block_caterpillar_ci_plot.rds")
block_ridge_plot                   <- readRDS("block_ridge_plot.rds")
block_savant_scatter_plot          <- readRDS("block_savant_scatter_plot.rds")
block_caterpillar_pages            <- readRDS("block_caterpillar_pages.rds")
block_ridge_pages                  <- readRDS("block_ridge_pages.rds")
```

# Abstract

Catcher defense is among the most difficult contributions to measure in baseball. We develop a hierarchical modeling framework isolating each catcher's defensive contribution across three dimensions: pitch framing, pitch blocking, and context dependent framing, applying mixed-effects logistic regression models to 2025 MLB Statcast data. Catcher random effects confirm that catcher identity explains meaningful variation in both borderline pitch calls and recorded block rates above and beyond pitch location, count, and pitcher tendencies. The framing model yields a 35-run gap between the best and worst framers, with Patrick Bailey leading at 21.8 runs saved; the blocking model finds a ~7 BAA/100 spread, with pitcher tendencies explaining more variation than catcher differences alone. Context dependent framing reveals that framing skill is pitch type specific: the main catcher random effect collapses to zero once pitch type interactions are included, implying that single-number framing leaderboards mask substantial within-catcher heterogeneity.

\newpage

# Introduction

In baseball, catchers play a pivotal role on defense, yet their value is historically difficult to measure. Traditional statistics such as passed balls and caught stealing rate capture only a fraction of what a catcher contributes. In this project we take a deep dive into three views of catcher defense: pitch framing, pitch blocking, and context dependent framing, quantifying each in terms of runs saved.

Pitch framing, the ability to influence umpires to call borderline pitches strikes, is one of the most impactful and measurable aspects of catcher defense. Elite framers can provide significant value, in the 2024 MLB season, Patrick Bailey saved 22 runs through framing alone (MLB Advanced Media, 2025). With the rise of Statcast pitch tracking, framing can now be rigorously isolated from pitch location and game context.

To measure framing, we fit a mixed-effects logistic regression model to 168,709 shadow zone pitches from the 2025 MLB season. Results reveal meaningful variation in framing ability with Patrick Bailey again leading the league with 21.8 runs saved, while Logan O'Hoppe and Shea Langeliers rank furthest below average at -13.3 and -13.0 runs respectively.

Blocking low and off-plate pitches is an important defensive trait for catchers, especially with runners on base where a missed block can allow immediate runs or runner advancement. Traditionally, Baseball Savant has used advanced inputs such as catcher positioning and called pitch type, which are not available in public pitch-level data, to build catcher blocking metrics and rankings (MLB Advanced Media, 2025). We aim to develop a comparable model for ranking catcher blocking ability using only publicly available pitch-level data.

Across all three analyses, catcher identity explains meaningful variation in defensive outcomes above and beyond contextual controls. The finding that the main catcher framing effect collapses entirely once pitch type interactions are included suggests that single-number framing leaderboards mask substantial within-catcher heterogeneity in defensive skill.

# Data

Pitch-level Statcast data was accessed via Baseball Savant's CSV export API for the 2025 MLB regular season (March 27 – September 28, 2025), yielding 712,192 total pitches across 118 variables (MLB Advanced Media, 2025). Key variables include `plate_x` and `plate_z` (pitch coordinates at home plate), `sz_top` and `sz_bot` (batter-specific strike zone boundaries), `fielder_2` (catcher ID), `description` (called strike or ball), `balls`, `strikes`, `stand`, and `p_throws`. Pitch-level coordinate data is uniquely suited for framing analysis as it allows precise separation of where the pitch actually crossed the plate from how the catcher influenced the umpire's call.

## Preprocessing: Catcher Framing

The dataset was filtered to called pitches only. Swings, fouls, and balls in play were excluded since framing only applies to pitches the umpire must judge. Pitches with missing plate coordinates, strike zone boundaries, or catcher ID were removed. A shadow zone filter was then applied, retaining only pitches within one baseball diameter of the strike zone edge, consistent with MLB's official shadow zone definition (MLB.com Glossary). The threshold of 0.242 ft was derived from the 2026 Official Baseball Rules circumference specification of 9.000–9.250 inches. Vertical pitch location was normalized (`pz_norm`) relative to each batter's individual strike zone to control for batter height differences. Finally, a minimum of 300 shadow zone pitches per catcher was applied to ensure reliable random effect estimates, reducing the dataset from 111 to 85 catchers and from 171,304 to 168,709 pitches. The final model dataset contains 168,709 shadow zone pitches across 85 catchers.

## Preprocessing: Catcher Blocking

We filtered the data to pitches with non-missing values for plate_x, plate_z, sz_top, sz_bot, fielder_2, pitch_type, and release_speed. Blocking opportunities were defined as below-zone pitches (plate_z \< sz_bot). The binary outcome is_block was set to 1 for blocked_ball or swinging_strike_blocked descriptions and 0 otherwise. Count was derived from the balls-strikes combination, and pitch types with fewer than 500 below-zone pitches were collapsed into an Other category. A minimum threshold of 100 below-zone pitches per catcher was applied, yielding a final dataset of 89,134 pitches across 88 catchers and 665 pitchers.

## Preprocessing: Context Dependent Framing

The shadow zone preprocessing follows the framing pipeline described in Section [Preprocessing: Catcher Framing], with two modifications. First, the column selection retains pitch_type so that each pitch can be assigned to a pitch family. Second, raw pitch type codes are grouped into six families (Fastball, Slider, Curveball, Changeup, Cutter, Splitter) plus an Other category, with sparse codes pooled to maintain stable per cell sample sizes. The final modeling dataset contains 168,709 shadow zone pitches across 85 catchers, identical to the framing model dataset.

## Exploratory Data Analysis: Catcher Framing

The shadow zone strike rate of 34.2% confirms that borderline pitches are genuinely ambiguous. A heatmap of called strike rate by location reveals a clear gradient from near-certain strikes at the interior edge to near-certain balls at the exterior edge (@fig-heatmap). Pitch volume per catcher is right-skewed with a long right tail of starting catchers, justifying the 300-pitch minimum cutoff (@fig-pitchdist). Called strike rate varies dramatically by count, from 57.0% on 3-0 counts to 10.6% on 0-2 counts, justifying inclusion of `count` as a fixed effect (@fig-countrate). Right-handed batters consistently receive higher called strike rates across nearly every count, justifying inclusion of `stand` as a fixed effect (@fig-handedness).

## Exploratory Data Analysis: Catcher Blocking

Statcast includes explicit labels for successful blocks through blocked_ball and swinging_strike_blocked. A heatmap of recorded block rate by pitch location confirms that block rates rise as pitches drop further below and further from the center of the zone (@fig-block-heatmap). Official blocking failures are far sparser: only 86 wild pitches and 29 passed balls appear in the des field, indicating that WP/PB outcomes substantially undercount true blocking failures and cannot serve as the primary outcome; recorded block rate is used instead. Block rate is nearly flat for pitches just below the zone and rises steeply beyond 0.3 ft, motivating a tighter depth filter (@fig-block-depth). Blocking opportunities are distributed unevenly across catchers, so a 100-pitch cutoff helps stabilize random-effect estimates (@fig-block-pitchdist). There is also substantial variation in block rate across pitch types, which is why sparse pitch types are collapsed into an Other category before model fitting (@fig-block-pitchtype).

## Exploratory Data Analysis: Context Dependent Framing

Before fitting interaction models, we examined the raw heterogeneity in shadow zone outcomes across count and pitch type jointly. Figure (@fig-eda-heatmap) displays called strike rate as a heatmap with count on the horizontal axis and pitch type on the vertical axis. Two patterns emerge. The horizontal gradient confirms count strongly shapes umpire behavior on borderline pitches, with 3-0 counts reaching 58 percent strike rates on fastballs and 0-2 counts dropping to 5 percent on splitters. More importantly, the vertical gradient within counts shows that pitch type also matters substantially, with curveballs and sliders consistently receiving higher strike rates than fastballs, changeups, and splitters in identical counts. The non-uniform pattern across the heatmap motivates the inclusion of catcher-by-count and catcher-by-pitch-type interaction terms in the modeling section.

# Methods

## Morgan Corliss: Catcher Framing

The goal of this analysis is to isolate each catcher's influence on borderline pitch calls above and beyond pitch location, count, and batter/pitcher context. We fit a mixed-effects logistic regression model where the binary outcome is whether a shadow zone pitch was called a strike (1) or ball (0):

$$\log\left(\frac{p_i}{1-p_i}\right) = \beta_0 + \beta_1 \text{plate\_x}_i + 
\beta_2 \text{pz\_norm}_i + \beta_3 \text{count}_i + \beta_4 \text{stand}_i + 
\beta_5 \text{p\_throws}_i + u_{\text{catcher}[i]} + u_{\text{pitcher}[i]} + 
u_{\text{batter}[i]}$$

where $p_i$ is the probability of a called strike on pitch $i$. Fixed effects control for pitch location, count, batter handedness, and pitcher handedness. The catcher random effect $u_{\text{catcher}[i]}$ is the primary quantity of interest, representing each catcher's tendency to influence umpire calls above or below league average after accounting for all other factors.

The model assumes $Y_i \sim \text{Bernoulli}(p_i)$, where the log-odds of a called strike are modeled as a linear function of fixed and random effects. Catcher, pitcher, and batter random effects are each assumed independently normally distributed with mean zero: $u_j \sim N(0, \sigma^2)$, independent of each other and of the fixed effect predictors. Shadow zone pitches are assumed conditionally independent given the random effects.

A mixed-effects model was chosen over fixed effects because with 85 catchers, a fixed-effects approach would produce unreliable estimates for backup catchers with small sample sizes. The mixed-effects framework handles this through partial pooling, shrinking estimates for data-sparse catchers toward the league mean. Pitcher and batter random effects were included to avoid attributing their tendencies to catchers. The BOBYQA optimizer was specified for this model due to its large dataset size and complex random effect structure (Powell, 2009).

To evaluate the model, we fit a reduced specification excluding pitcher and batter random effects:

$$\text{is\_strike} \sim \text{plate\_x} + \text{pz\_norm} + \text{count} + 
\text{stand} + \text{p\_throws} + (1 | \text{catcher\_id})$$

The full and reduced models were compared via AIC and a likelihood ratio test to assess whether adding pitcher and batter random effects significantly improves fit.

Uncertainty in catcher random effect estimates is quantified using conditional standard errors from the mixed-effects model. Ninety percent confidence intervals are computed as estimate ± 1.645 standard errors, reflecting uncertainty under the assumption that the model is correctly specified.

## Arthur Spirou: Catcher Blocking via Mixed-Effects Logistic Regression

I fit a mixed-effects logistic regression model where the binary outcome is whether a below-zone pitch resulted in a recorded block (1) or not (0), isolating each catcher’s influence above and beyond pitch location, pitch type, and pitcher tendencies:

$$\log\left(\frac{p_i}{1-p_i}\right) = \beta_0 + \beta_1\,\text{plate\_x}_i +
\beta_2\,\text{plate\_z}_i + \boldsymbol{\gamma}^{\top}\mathbf{z}_i +
\beta_3\,\text{release\_speed}_i + \beta_4\,\text{count}_i +
\beta_5\,\text{p\_throws}_i + u_{j(i)} + v_{k(i)}$$

where $p_i$ is the probability of a recorded block on pitch $i$. Fixed effects control for horizontal and vertical pitch location, collapsed pitch type ($\mathbf{z}_i$), release speed, count, and pitcher handedness. The catcher random effect $u_{j(i)}$ is the primary quantity of interest, representing each catcher’s tendency to record blocks above or below league average. The pitcher random effect $v_{k(i)}$ controls for pitcher-level differences in block-inducing pitch tendencies. Catcher and pitcher random effects are assumed independently $N(0, \sigma^2_\text{catcher})$ and $N(0, \sigma^2_\text{pitcher})$, with pitches conditionally independent given the random effects.

A mixed-effects model was chosen for the same reasons as the framing model: partial pooling shrinks sparse-catcher estimates toward the league mean, avoiding unreliable fixed-effect estimates for low-volume catchers. Pitcher random effects control for pitcher-level variation in below-zone pitch frequency and difficulty; omitting them would attribute that variation to catchers. The BOBYQA optimizer was specified for the large dataset and crossed random effect structure (Powell, 2009).

I compared a simple specification (pitch location and type only) against a full specification additionally including release speed, count, and pitcher handedness:

$$\text{Simple:}\quad \text{is\_block} \sim \text{plate\_x} + \text{plate\_z} +
\text{pitch\_type} + (1 \mid \text{catcher\_id}) + (1 \mid \text{pitcher\_id})$$

$$\text{Full:}\quad +\;\text{release\_speed} + \text{count} + \text{p\_throws}$$

The full model achieved substantially lower AIC (61,287.6 vs. 61,972.9, $\Delta$AIC = 685.3; $\chi^2$(13) = 711.29, p < 2.2e-16) and was selected for all subsequent inference (@tbl-block-comparison).

Catcher random effects are expressed as blocks above average per 100 opportunities (BAA/100). Let $\hat{\beta}_0$ denote the model intercept and $\hat{u}_j$ the estimated random effect for catcher $j$. The league baseline block probability is
$\hat{p}_\text{base} = \text{logistic}(\hat{\beta}_0)$ and the
catcher-specific probability is $\hat{p}_j = \text{logistic}(\hat{\beta}_0
+ \hat{u}_j)$. Then $\text{BAA/100}_j = (\hat{p}_j - \hat{p}_\text{base})
\times 100$.

Uncertainty is quantified using conditional standard errors via `arm::se.ranef()`, approximating the Laplace-approximated posterior curvature for each catcher BLUP. Ninety percent confidence intervals are $\hat{u}_j \pm 1.645 \times \text{SE}_j$, transformed to the BAA/100 scale. Parametric bootstrap via `bootMer()` was computationally infeasible at approximately 422 seconds per iteration (over 23 hours for B = 200); conditional standard errors treat the estimated variance components as fixed, introducing slight undercoverage for data-sparse catchers.

## Joel Maison: Context Dependent Catcher Framing

Standard framing analyses estimate a single defensive value per catcher and treat framing ability as constant across all game contexts. This subsection asks a structurally different question: is framing skill itself uniform, or does each catcher's ability to influence borderline pitch calls depend on the context in which the pitch is thrown? Two contextual dimensions are tested: the count and the pitch type.

To test count dependence, the framing specification is extended with a catcher by count interaction random effect. To test pitch type dependence, a separate extension adds a catcher by pitch type interaction. Both extensions share the same fixed effect structure described in the framing methods and operate on the same shadow zone dataset of 168,709 pitches across 85 catchers.

The count interaction model is specified as:

$$\log\left(\frac{p_i}{1-p_i}\right) = \beta_0 + \beta_1 \text{plate\_x}_i + 
\beta_2 \text{pz\_norm}_i + \beta_3 \text{count}_i + \beta_4 \text{stand}_i + 
\beta_5 \text{p\_throws}_i + u_{\text{catcher}[i]} + u_{\text{catcher}[i] \times \text{count}[i]} + 
u_{\text{pitcher}[i]} + u_{\text{batter}[i]}$$

The pitch type interaction model replaces the catcher by count term with a catcher by pitch type term:

$$\log\left(\frac{p_i}{1-p_i}\right) = \beta_0 + \beta_1 \text{plate\_x}_i + 
\beta_2 \text{pz\_norm}_i + \beta_3 \text{count}_i + \beta_4 \text{stand}_i + 
\beta_5 \text{p\_throws}_i + u_{\text{catcher}[i]} + u_{\text{catcher}[i] \times \text{pitchtype}[i]} + 
u_{\text{pitcher}[i]} + u_{\text{batter}[i]}$$

In each model, the interaction term captures each catcher's deviation from their own average framing ability within a specific context (count or pitch type) and is the primary quantity of interest. The remaining catcher, pitcher, and batter random intercepts control for overall tendencies of those actors.

The binary outcome follows a Bernoulli distribution, $Y_i \sim
\text{Bernoulli}(p_i)$, with the log-odds modeled as a linear function of fixed and random effects. All random effects are assumed independently normally distributed with mean zero, $u_j \sim N(0,
\sigma^2)$, independent of each other and of the fixed effect predictors. Shadow zone pitches are assumed conditionally independent given the random effects.

The interactions are modeled as random effects rather than fixed effects to enable partial pooling. A fixed effect specification would require 1,020 parameters for the catcher by count interaction and 595 parameters for the catcher by pitch type interaction, producing unstable estimates for catchers with limited data in rare cells. The random effect framework pools information across all cells, producing stable estimates even for sparse combinations. The BOBYQA optimizer is used for all model fits, following Powell (2009).

To evaluate whether each interaction earns its keep, two likelihood ratio tests are conducted. The count interaction model is tested against a reduced specification with no catcher by count term, fit on the same 168,709 row dataset. The pitch type interaction model is tested against an analogously reduced specification fit on the 165,424 row subset where pitch type is non missing. AIC values are reported alongside chi squared statistics and p values to characterize the practical and statistical significance of each extension.

Uncertainty in random effect estimates is quantified using conditional standard errors from the fitted glmer objects. Ninety percent confidence intervals are computed as estimate plus or minus 1.645 standard errors, reflecting uncertainty under the assumption of correct model specification.

# Results

## Morgan Corliss: Catcher Framing

### Model Comparison

The full model achieved a substantially lower AIC (204,945) compared to the reduced model (205,241), a difference of 296 units. The likelihood ratio test confirms this improvement is statistically significant (χ²(2) = 300.25, p \< 2.2e-16), providing strong evidence that pitcher and batter random effects meaningfully improve model fit. Omitting them would bias catcher framing estimates by attributing pitcher and batter tendencies to catchers. The full model was therefore selected for all subsequent inference. These results are summarized in @tbl-model-comparison.

### Model Interpretation

@tbl-variance-fixed displays the variance components and key fixed effects. Catcher variance (0.006) is smaller than pitcher (0.013) and batter (0.016) variance, suggesting that pitch location and batter/pitcher context explain more variation in borderline calls than catcher framing alone. However, catcher variance is meaningfully non-zero, confirming that framing ability varies across catchers above and beyond these controls, directly supporting the motivation for this analysis.

Pitcher handedness was not significant (β = -0.012, p = 0.468) and was retained for completeness. Count coefficients were highly significant (p \< 2e-16), with 3-0 counts showing the highest tendency toward called strikes (β = 0.586) and 0-2 counts the lowest (β = -1.897). Right-handed batters received significantly more called strikes than left-handed batters (β = 0.140, p \< 2e-16), consistent with @fig-handedness.

### Catcher Framing Estimates

@fig-caterpillar displays the framing random effects for all 85 catchers with 90% confidence intervals. Roughly 15 catchers are significantly different from zero, confirming that elite and poor framers are reliably distinguishable from average while many mid-tier catchers have intervals that overlap zero.

@fig-runssaved and @tbl-leaderboard show the top and bottom 10 catchers by total runs saved. These results directly address the motivating question of how much catcher framing contributes to defensive value, the gap between the best and worst framers is approximately 35 runs over a full season, a difference large enough to meaningfully impact team performance. Patrick Bailey leads all catchers with 21.8 runs saved (90% CI: 13.9 to 29.6), consistent with his reputation as the best framer in baseball and with the 22 runs cited in the introduction for the 2024 season. Logan O'Hoppe (-13.3 runs, CI: -20.5 to -6.2) and Shea Langeliers (-13.0 runs, CI: -20.6 to -5.5) rank furthest below average. Austin Hedges saves the most runs per 100 shadow zone pitches (0.49) despite limited playing time, suggesting strong framing skill regardless of sample size. Notable catchers with confidence intervals crossing zero including Austin Wells and Carlos Narváez cannot be confidently distinguished from average, highlighting the importance of reporting uncertainty alongside point estimates.

## Arthur Spirou: Catcher Blocking Results

### Model Comparison

The full model achieved lower AIC (61,287.6 vs. 61,972.9, ΔAIC = 685.3; χ²(13) = 711.29, p < 2.2e-16), confirming that release speed, count, and pitcher handedness improve fit beyond pitch location and type alone (@tbl-block-comparison).

### Model Interpretation

Pitcher-level variance (0.062) exceeds catcher-level variance (0.035), indicating that pitcher tendencies explain more variation in recorded block rate than catcher differences alone; non-zero catcher variance confirms that blocking ability varies meaningfully above and beyond pitch difficulty. Full variance components and fixed effects are in @tbl-block-fixed.

### Catcher Blocking Estimates

@fig-block-caterpillar shows BAA/100 estimates for all 88 catchers with 90% confidence intervals. Patrick Bailey leads at +4.04 BAA/100 and Agustín Ramírez ranks last at −2.92, a spread of approximately 7 units indicating that catcher identity produces practically meaningful differences in blocking outcomes.

The median 90% CI width is approximately 3.33 BAA/100, with most catchers clustering near zero with overlapping intervals; only those whose interval excludes zero are reliably distinguished from league average. @fig-block-ridge displays the model-implied distributions N(û_j, SE²_j) for each catcher. Full rankings are in @tbl-block-summary.

### External Validation

I compared my rankings against Baseball Savant's 2025 blocking leaderboard (MLB Advanced Media, 2025) via Spearman rank correlation across 65 matched catchers. The correlation was ρ = −0.072 (@fig-block-savant-scatter), reflecting structural differences rather than model failure: my outcome is recorded block rate on pitches 0.3–1.5 ft below zone, while Savant measures WP/PB prevention weighted by pre-pitch probability across a broader population (@tbl-block-opps).

Alejandro Kirk illustrates this: Savant's top-ranked blocker in 2025 sits at −0.37 BAA/100 (rank 55 of 88) in my model. His elite positioning likely converts would-be wild pitches into clean receives that never receive the `blocked_ball` label, invisible to my outcome. My model rewards the act of blocking; Savant rewards WP/PB prevention achievable through pre-pitch positioning. These are genuinely distinct skills.

## Joel Maison: Context Dependent Catcher Framing

### Likelihood Ratio Tests

Both interaction extensions improve model fit relative to the
specifications without them, but at very different magnitudes. The
count interaction model achieves an AIC of 204,937.5 compared to
204,944.9 for the model without the catcher by count term, a
difference of 7.4 units. The likelihood ratio test confirms this
improvement is statistically significant ($\chi^2(1) = 9.46$, $p =
0.002$). The pitch type interaction model achieves a substantially
larger improvement: AIC drops from 201,012 to 200,702, a difference of
310 units, with the likelihood ratio test returning $\chi^2(1) =
312.26$, $p < 2.2e-16$. The pitch type interaction is therefore an
order of magnitude more important than the count interaction in
explaining shadow zone outcomes. These results are summarized in @tbl-lrt-summary.

### Variance Components

The variance components reveal the structural finding of this
subsection. In the count interaction model, the catcher by count
random effect has a standard deviation of 0.062, smaller than the main
catcher effect (0.075), suggesting that count specific framing
deviations are real but secondary to a catcher's overall framing
skill. In the pitch type interaction model, the catcher by pitch type
random effect has a standard deviation of 0.191, more than double the
main catcher effect in the count model. More striking, the main
catcher random effect in the pitch type model collapses to zero
(standard deviation 0.000). Once each catcher is permitted a different
framing skill per pitch type, no general framing skill remains to
estimate. Variance components for both models are reported in @tbl-variance-components.

This collapse has a substantive interpretation. Catcher framing is not
one skill that varies modestly by context. It is a collection of pitch
type specific skills, and a single number per catcher framing
estimate, such as the leaderboard reported in the framing section,
implicitly averages across these per pitch type abilities.

### Catcher Specialization Patterns

@fig-pt-topbot displays the top and bottom five catchers by runs
saved per 100 pitches in the four most common pitch type categories
(fastball, slider, curveball, changeup). The composition of the top
five differs across pitch types. Yohel Pozo, Austin Hedges, and Pedro
Pagés lead on fastballs but do not appear among the top five
curveball framers. Patrick Bailey, who leads the overall framing
leaderboard reported earlier in this report, appears in the top five
on fastballs and curveballs but not on sliders or changeups. Hayden
Senger leads on sliders but is not visible in the other panels. The
non overlapping membership of these top five lists is the visual
expression of the variance collapse described above.

@fig-pt-profile shows full pitch type profiles for the six catchers
with the largest spread in framing runs across pitch types. A
recurring pattern emerges. Several catchers, including J.T. Realmuto,
Sean Murphy, and Drake Baldwin, post positive framing values on
fastballs, sliders, and curveballs while posting clearly negative
values on changeups and splitters. The hard pitch versus offspeed
divide accounts for much of the within catcher variation visible in
the figure. Other catchers, such as Keibert Ruiz, show the inverse
pattern, performing better on offspeed pitches than on the harder
arsenal.

### Catcher Baseline Effects

The catcher baseline random effects from the count interaction model
are presented in @fig-main-caterpillar. Approximately 15 of the 85
catchers have 90 percent confidence intervals that exclude zero,
indicating reliably above or below average baseline framing ability
under that model. The remaining catchers cluster near zero with
intervals overlapping the league average, consistent with limited
shadow zone exposure for many backup catchers in a single season.

# Discussion

## Conclusions

This project examined three key aspects of catcher defense: pitch
framing, pitch blocking, and context dependent framing, quantifying
each in terms of runs saved. For framing, the mixed-effects model
confirms that catcher identity explains meaningful variation in
borderline pitch calls above and beyond pitch location, count,
pitcher, and batter effects. The gap between the best and worst
framers is approximately 35 runs over a full season, a difference
large enough to meaningfully impact team performance and roster
decisions. Patrick Bailey's 21.8 runs saved leads all catchers,
consistent with his standing as the best framer in baseball.

The blocking model confirms that catcher identity explains meaningful
variation in recorded block rates above and beyond pitch difficulty.
Patrick Bailey again leads at +4.04 blocks above average per 100
opportunities, while Agustín Ramírez ranks last at −2.92, a spread of
roughly 7 units across 88 catchers. Notably, pitcher-level variance
(0.062) exceeds catcher-level variance (0.035), indicating that pitcher
tendencies to miss below the zone explain more variation in recorded
block rate than catcher differences alone.

The context dependent framing analysis qualifies this single number
view of catcher framing skill. The catcher by count interaction is
statistically significant ($\chi^2(1) = 9.46$, $p = 0.002$) but
modest in magnitude. The catcher by pitch type interaction is far
more substantial ($\chi^2(1) = 312.26$, $p < 2.2e-16$), and once it
is included the main catcher random effect collapses to zero. This
implies that catcher framing is fundamentally pitch type specific
rather than a single general skill, and that the framing leaderboard
above implicitly averages over pitch type specific abilities that
differ substantially across catchers.

## Limitations

Several limitations of this analysis should be noted. First, the shadow zone threshold of 0.242 ft is derived from the average baseball diameter per the 2026 Official Baseball Rules and a robustness check with alternative values would strengthen this analysis. Second, this analysis is based on a single season of data, framing estimates from one season may reflect some noise. Conducting an analysis of different years would help verify our conclusions. Uncertainty is quantified via conditional standard errors, which assume the model is correctly specified, a bootstrap approach would provide more robust intervals but was proving computationally infeasible.

The blocking analysis has three key limitations. First, is_block = 0 conflates clean receives with missed blocks: official WP/PB outcomes are too sparse to distinguish them reliably, and many mishandled pitches are never scored as official failures. Second, the below-zone filter is a practical approximation: it may exclude some horizontally difficult pitches and may include low pitches that were easy to receive. Third, the model is limited to publicly available pitch-level variables and cannot incorporate catcher starting position, body orientation, or other tracking inputs used in proprietary models like Baseball Savant's.

The context dependent framing analysis has two notable limitations. First, while the catcher by pitch type interaction is statistically overwhelming, individual per cell estimates remain noisy because sample sizes per catcher per pitch type are uneven. Aggregate variance components are reliable but specific catcher rankings within pitch types should be interpreted with caution. Second, uncertainty quantification relies on conditional standard errors, which assume correct model specification, rather than a bootstrap approach that proved computationally infeasible within the project timeline.

## Future Work

Several extensions of this work would strengthen the conclusions. A year-to-year stability check using 2024 data would test whether catcher rankings are consistent across seasons; high correlation would confirm the models measure real defensive skill rather than single-season noise, and the same check applied to the pitch type interaction would test whether catcher specialization patterns are stable. Cross validation alongside AIC and likelihood ratio tests would provide additional evidence on out-of-sample performance. A throwing component based on stolen base data retrieved from a separate Statcast endpoint would complete the three-pillar defensive framework.

\newpage

# References

MLB.com. "Catcher Framing." *MLB Glossary*, Major League Baseball, https://www.mlb.com/glossary/statcast/catcher-framing. Accessed April 2026.

MLB Advanced Media. "Catcher Framing Leaderboard." *Baseball Savant*, Major League Baseball, 2025. https://baseballsavant.mlb.com/catcher_framing. Accessed April 2026.

MLB Advanced Media. "Catcher Blocking Leaderboard." *Baseball Savant*, Major League Baseball, 2025. https://baseballsavant.mlb.com/leaderboard/catcher-blocking. Accessed April 2026.

MLB Advanced Media. *Statcast Pitch-Level Data, 2025 MLB Regular Season.* Baseball Savant, Major League Baseball, 2025. https://baseballsavant.mlb.com/statcast_search. Accessed April 2026.

Major League Baseball. *2026 Official Baseball Rules*, pg. 5. https://mktg.mlbstatic.com/mlb/official-information/2026-official-baseball-rules.pdf. Accessed April 2026.

Powell, M.J.D. "The BOBYQA Algorithm for Bound Constrained Optimization Without Derivatives." *Cambridge NA Report NA2009/06*, University of Cambridge, 2009. https://www.damtp.cam.ac.uk/user/na/NA_papers/NA2009_06.pdf. Accessed April 2026.

## AI Usage Statement

Arthur used Claude Code in VS Code to generate blocking analysis code based on my model description and methods from the rough draft, an outline of the blocking analysis work to help structure my section's report content, and to merge file, outputs, and analysis structures for cohesiveness with my teammates. At all points I made the assumptions, choices, and interpretations myself based upon course materials, my own knowledge, and supporting research, while utilizing Claude Code to implement the designs. Because this was accomplished in VS Code only, I do not have a shareable conversation history link, so this is a summary of my AI usage to the best of my ability.

\newpage

# Appendix

Code repository: <https://github.com/mcorliss7239/sports-analytics-catcher-contributions>

## Tables

```{r}
#| echo: false
#| label: tbl-model-comparison
#| tbl-cap: "Model comparison: full vs. reduced mixed-effects logistic regression"

model_comparison_table
```

```{r}
#| echo: false
#| label: tbl-variance-fixed
#| tbl-cap: "Variance components and key fixed effects from full mixed-effects model"

variance_fixed_table

```

```{r}
#| echo: false
#| label: tbl-count-comparison
#| tbl-cap: "Model comparison: baseline framing model vs. count-specific interaction model"

tibble(
  Model         = c("Baseline", "Count interaction"),
  `RE Structure` = c(
    "Catcher + Pitcher + Batter",
    "+ Catcher x Count"
  ),
  AIC           = c(204944.9, 204937.5),
  `Delta AIC`   = c(7.4, 0.0),
  `Chi-sq`      = c("—", "9.457"),
  `df`          = c("—", "1"),
  `p-value`     = c("—", "0.002")
) %>%
  kable(
    format = "simple",
    digits = 1
  )
```

\newpage
```{r}
#| echo: false
#| label: tbl-leaderboard
#| tbl-cap: "Top and bottom 10 catchers by framing runs saved (2025 season)"
leaderboard_table
```

\newpage

```{r}
#| echo: false
#| label: tbl-block-comparison
#| tbl-cap: "Blocking model comparison: simple vs. full mixed-effects logistic regression"

{
  tbl <- block_model_comparison_table
  attr(tbl, "caption") <- NULL
  tbl <- gsub("\\\\caption\\{(?:[^{}]|\\{[^}]*\\})*\\}\\n?", "", tbl, perl = TRUE)
  tbl
}
```

```{r}
#| echo: false
#| label: tbl-block-fixed
#| tbl-cap: "Variance components and fixed effects from the full blocking model. Reference categories: pitch type = Changeup (CH), count = 0-0, pitcher handedness = left-handed."

{
  tbl <- block_variance_fixed_table
  attr(tbl, "caption") <- NULL
  tbl <- gsub("\\\\caption\\{(?:[^{}]|\\{[^}]*\\})*\\}\\n?", "", tbl, perl = TRUE)
  tbl
}
```

```{r}
#| echo: false
#| label: tbl-block-summary
#| tbl-cap: "Catcher blocking rankings: 2025 season (minimum 100 below-zone pitches, 0.3–1.5 ft below zone)"

catcher_effects_named |>
  dplyr::arrange(desc(baa_per_100)) |>
  dplyr::select(label, n_opps, baa_per_100) |>
  dplyr::mutate(baa_per_100 = round(baa_per_100, 3)) |>
  knitr::kable(col.names = c("Catcher", "Blocking Opportunities", "Blocks Above Avg / 100 Opp"))
```

```{r}
#| echo: false
#| label: tbl-block-opps
#| tbl-cap: "Blocking opportunity counts by system (2025 season, matched catchers). Our opportunities are pitches 0.3–1.5 ft below the batter-specific zone bottom; Savant pitches are total pitches received."

{
  tbl <- block_opportunity_comparison_table
  attr(tbl, "caption") <- NULL
  tbl <- gsub("\\\\caption\\{(?:[^{}]|\\{[^}]*\\})*\\}\\n?", "", tbl, perl = TRUE)
  tbl
}
```

\newpage

```{r}
#| echo: false
#| label: tbl-lrt-summary
#| tbl-cap: "Likelihood ratio tests for context dependent framing extensions"
{
  tbl <- lrt_summary
  attr(tbl, "caption") <- NULL
  tbl <- gsub("\\\\caption\\{(?:[^{}]|\\{[^}]*\\})*\\}\\n?", "", tbl, perl = TRUE)
  tbl
}
```

\newpage
```{r}
#| echo: false
#| label: tbl-variance-components
#| tbl-cap: "Variance components from the count and pitch type interaction models"
{
  tbl <- variance_components_table
  attr(tbl, "caption") <- NULL
  tbl <- gsub("\\\\caption\\{(?:[^{}]|\\{[^}]*\\})*\\}\\n?", "", tbl, perl = TRUE)
  tbl
}
```



## Figures

```{r}
#| echo: false
#| warning: false
#| fig-cap: "Called strike rate by pitch location (2025 shadow zone). Solid box = strike zone, dashed box = shadow zone boundary. Interior of strike zone excluded by design."
#| label: fig-heatmap
#| fig-width: 6
#| fig-height: 5

heatmap_plot
```

```{r}
#| echo: false
#| warning: false
#| fig-cap: "Distribution of shadow zone pitches per catcher (2025 season). Dashed line indicates 300-pitch minimum threshold."
#| label: fig-pitchdist
#| fig-width: 6
#| fig-height: 4

pitchdist_plot
```

```{r}
#| echo: false
#| warning: false
#| fig-cap: "Called strike rate by count in shadow zone (2025 season)."
#| label: fig-countrate
#| fig-width: 6
#| fig-height: 5

countrate_plot
```

```{r}
#| echo: false
#| warning: false
#| fig-cap: "Called strike rate by count and batter handedness (2025 shadow zone)."
#| label: fig-handedness
#| fig-width: 6
#| fig-height: 5

handedness_plot
```

```{r}
#| echo: false
#| warning: false
#| label: fig-caterpillar
#| fig-cap: "Catcher framing random effects with 90% confidence intervals (2025 season). Red points indicate catchers significantly different from zero."
#| fig-width: 7
#| fig-height: 6

caterpillar_plot
```

```{r}
#| echo: false
#| warning: false
#| label: fig-runssaved
#| fig-cap: "Catcher framing value for top and bottom 10 catchers (2025 season). Error bars show 90% confidence intervals."
#| fig-width: 7
#| fig-height: 6

runssaved_plot
```

```{r}
#| echo: false
#| warning: false
#| label: fig-block-heatmap
#| fig-cap: "Recorded block rate by pitch location (2025 below-zone pitches). Dashed line = average strike zone bottom. Block rates rise as pitches drop lower and move further from center."
#| fig-width: 6
#| fig-height: 5

block_heatmap_plot
```

```{r}
#| echo: false
#| warning: false
#| label: fig-block-depth
#| fig-cap: "Recorded block rate by depth below zone (2025 season). Block rate is nearly flat for pitches 0–0.3 ft below the zone and rises steeply beyond, motivating the 0.3 ft opportunity filter."
#| fig-width: 6
#| fig-height: 4

block_by_depth_plot
```

```{r}
#| echo: false
#| warning: false
#| label: fig-block-pitchdist
#| fig-cap: "Below-zone pitches per catcher (2025 season). Dashed line = 100-pitch minimum threshold."
#| fig-width: 6
#| fig-height: 4

block_pitchdist_plot
```

```{r}
#| echo: false
#| warning: false
#| label: fig-block-pitchtype
#| fig-cap: "Recorded block rate by pitch type (2025 below-zone pitches). Breaking balls show higher block rates; sparse pitch types are collapsed into Other before modeling."
#| fig-width: 6
#| fig-height: 4

block_by_pitchtype_plot
```

\newpage

```{r}
#| echo: false
#| warning: false
#| label: fig-block-caterpillar
#| fig-cap: "Catcher blocking effects: blocks above average per 100 opportunities (2025 season), with 90% confidence intervals from conditional standard errors (Laplace approximation). Catchers whose interval excludes zero are reliably above or below average."
#| fig-width: 7
#| fig-height: 9

block_caterpillar_ci_plot
```

\newpage

```{r}
#| echo: false
#| warning: false
#| label: fig-block-ridge
#| fig-cap: "Model-implied blocking ability distributions (2025 season). Each curve is N(u-hat_j, SE_j^2) from the fitted model; width reflects estimation uncertainty."
#| fig-width: 7
#| fig-height: 9

block_ridge_plot
```

\newpage

```{r}
#| echo: false
#| warning: false
#| label: fig-block-savant-scatter
#| fig-cap: "Our blocking model vs. Baseball Savant (2025 season). Each point is a catcher present in both systems (n = 65). Spearman rho = -0.072, reflecting structural differences in opportunity definition and outcome between the two systems."
#| fig-width: 6
#| fig-height: 7

block_savant_scatter_plot
```

\newpage

```{r}
#| echo: false
#| warning: false
#| fig-cap: "Catcher blocking estimates (BAA/100) with 90% confidence intervals, page 1 of 5."
#| fig-width: 7
#| fig-height: 9
block_caterpillar_pages[[1]]
```

\newpage

```{r}
#| echo: false
#| warning: false
#| fig-cap: "Catcher blocking estimates (BAA/100) with 90% confidence intervals, page 2 of 5."
#| fig-width: 7
#| fig-height: 9
block_caterpillar_pages[[2]]
```

\newpage

```{r}
#| echo: false
#| warning: false
#| fig-cap: "Catcher blocking estimates (BAA/100) with 90% confidence intervals, page 3 of 5."
#| fig-width: 7
#| fig-height: 9
block_caterpillar_pages[[3]]
```

\newpage

```{r}
#| echo: false
#| warning: false
#| fig-cap: "Catcher blocking estimates (BAA/100) with 90% confidence intervals, page 4 of 5."
#| fig-width: 7
#| fig-height: 9
block_caterpillar_pages[[4]]
```

\newpage

```{r}
#| echo: false
#| warning: false
#| fig-cap: "Catcher blocking estimates (BAA/100) with 90% confidence intervals, page 5 of 5."
#| fig-width: 7
#| fig-height: 9
block_caterpillar_pages[[5]]
```

\newpage

```{r}
#| echo: false
#| warning: false
#| fig-cap: "Model-implied blocking ability distributions N(û_j, SE²_j), page 1 of 5."
#| fig-width: 7
#| fig-height: 9
block_ridge_pages[[1]]
```

\newpage

```{r}
#| echo: false
#| warning: false
#| fig-cap: "Model-implied blocking ability distributions N(û_j, SE²_j), page 2 of 5."
#| fig-width: 7
#| fig-height: 9
block_ridge_pages[[2]]
```

\newpage

```{r}
#| echo: false
#| warning: false
#| fig-cap: "Model-implied blocking ability distributions N(û_j, SE²_j), page 3 of 5."
#| fig-width: 7
#| fig-height: 9
block_ridge_pages[[3]]
```

\newpage

```{r}
#| echo: false
#| warning: false
#| fig-cap: "Model-implied blocking ability distributions N(û_j, SE²_j), page 4 of 5."
#| fig-width: 7
#| fig-height: 9
block_ridge_pages[[4]]
```

\newpage

```{r}
#| echo: false
#| warning: false
#| fig-cap: "Model-implied blocking ability distributions N(û_j, SE²_j), page 5 of 5."
#| fig-width: 7
#| fig-height: 9
block_ridge_pages[[5]]
```

\newpage

```{r}
#| echo: false
#| warning: false
#| label: fig-count-effects
#| fig-cap: "Top and bottom 15 catcher-by-count random effects (2025 season). Each point represents a catcher-count combination. Positive values indicate above-average framing in that count."
#| fig-width: 7
#| fig-height: 5

# library(tidyverse)
# library(lme4)
# 
# ranef(framing_count)$`catcher_id:count` %>%
#   rownames_to_column("catcher_count") %>%
#   rename(effect = `(Intercept)`) %>%
#   arrange(desc(effect)) %>%
#   slice(c(1:15, (n()-14):n())) %>%
#   mutate(
#     catcher_count = fct_reorder(catcher_count, effect),
#     direction = ifelse(effect > 0, "Above average", "Below average")
#   ) %>%
#   ggplot(aes(x = effect, y = catcher_count, color = direction)) +
#   geom_point(size = 2.5) +
#   geom_vline(xintercept = 0, linetype = "dashed", linewidth = 0.5) +
#   scale_color_manual(
#     values = c("Above average" = "steelblue", "Below average" = "firebrick"),
#     name = NULL
#   ) +
#   labs(
#     title    = "Catcher-by-count framing effects (2025 season)",
#     subtitle = "Top and bottom 15 catcher-count combinations",
#     x        = "Random effect (log-odds scale)",
#     y        = NULL
#   ) +
#   theme_bw() +
#   theme(
#     panel.grid.minor   = element_blank(),
#     panel.grid.major.y = element_blank(),
#     legend.position    = "bottom"
#   )
```

```{r}
#| echo: false
#| warning: false
#| label: fig-eda-heatmap
#| fig-cap: "Called strike rate by count and pitch type (2025 shadow zone). Cells show the percentage of shadow zone pitches called strikes for each combination."
framing_eda_count_pt_plot
```

```{r}
#| echo: false
#| warning: false
#| label: fig-main-caterpillar
#| fig-cap: "Catcher baseline framing effects from the count interaction model (2025 season). 90 percent confidence intervals from conditional standard errors. Red points indicate catchers significantly different from zero."
framing_main_plot
```

```{r}
#| echo: false
#| warning: false
#| label: fig-pt-topbot
#| fig-cap: "Top and bottom five catchers by runs saved per 100 shadow zone pitches, faceted by pitch type (fastball, slider, curveball, changeup). 90 percent confidence intervals from conditional standard errors."
framing_pt_topbot_plot
```

```{r}
#| echo: false
#| warning: false
#| label: fig-pt-profile
#| fig-cap: "Catcher framing profiles for the six catchers with the largest spread in runs saved across pitch types (2025 season). 90 percent confidence intervals from conditional standard errors."
framing_pt_profile_plot
```

